---
title: |
  | \vspace*{3cm} Fielding Lung WGS - SCLC Lung Cancer
  | DNA Methylation Analysis 
  |  \vspace{2cm}
subtitle: |
  | \vspace*{1cm} SCLC WGS Methylation and 
  | Association to Genes of Interest
  |  \vspace{3cm}
author: |
  | Vanessa Lakis \vspace{3cm} 
date: |
  | `r format(Sys.time(), "%d %B, %Y")` \newpage 
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_tex: yes
  html_document:
    toc: true
highlight: zenburn
  
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{xcolor, hyperref}
  - \usepackage{lmodern}
  - \usepackage{courier}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{pdflscape}  
  - \newcommand{\blandscape}{\begin{landscape}} 
  - \newcommand{\elandscape}{\end{landscape}}
  - \setlength{\headheight}{36.5pt}
  - \setlength{\footskip}{28pt}
  - \pagestyle{fancy}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \cfoot{\thepage}
  - \rhead{\scriptsize The University of Queensland \\ 280-284 Sir Fred Schonell Dr, St Lucia, QLD 4067 Australia\\T +61 (7) 3365 1111 | W \url{www.uq.edu.au}}
  - \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \usepackage{float}
---

```{r global-options, include=FALSE}
# **********
# load packages required - sets out the document rendering defaults and loads the R packages that are required for analysis
# **********
library(knitr)
knitr::opts_chunk$set(dpi=100, echo=FALSE, warning=FALSE, message=FALSE, 
                      fig.show='asis', fig.keep='all', fig.width=7, fig.height=4, fig.pos="H", fig.align='center') 
# to remove message from xtable
library(xtable)
options(xtable.comment=FALSE)
 
loadpak = function(pkg){
  sapply(pkg, require, character.only=TRUE)
}
# acquire library for the analysis
packages = c("dplyr", "data.table", "ggplot2", "ggfortify", "gtable", "sva", "scales", "ChAMP", "fpc", "cowplot",
             "gridExtra", "RColorBrewer", "RPMM", "reshape2", "pheatmap", "lumi", "circlize", "readr", "readxl",
             "stringi", "dendextend", "preprocessCore", "ComplexHeatmap", "ReactomePA", "magrittr", "ggpubr", "psych", "xlsx") 
loadpak(packages)
# packages used for the pathway analysis
packages = c("AnnotationDbi", "org.Hs.eg.db", "clusterProfiler", "DOSE", "msigdbr", "fgsea", "enrichplot", "pathview")
loadpak(packages)

```


```{r study-options, include=FALSE, results='asis'}
# **********
# import functions for script
# *********
#
#   **********************  nominate study directory ***************************
#
study_dir = ""
source(paste0(study_dir, "Methylation/WGS_SCLC_Analysis/Fielding_-_Lung_SCLC_Meth_Functions_Uplaod.R"))

# **********
# import variable declarations for script
# *********
source(paste0(study_dir, "Methylation/WGS_SCLC_Analysis/Fielding_-_Lung_SCLC_Meth_Declarations_Upload.R"))
```  


```{r read-in-rdata-files}
# **********
# loads in the data imported and normalised from idat files.  If you wish to used results from previous analysis then make first_run=FALSE and normalise=FALSE
# **********
if (!first_run) {
  RData_datafile = paste0(study_prefix, "_", gsub("/Differential_stats", "", analysis), ifelse(nchar(sub_analysis)>12, paste0("_", sub_analysis), "" ), "_Raw.RData")
  if (RData_datafile %in% dir(RData_results_dir)) {
    load(paste0(RData_results_dir, RData_datafile))
  } else {
    first_run = TRUE
  }
}
if (!normalise) {
  RData_normfile = paste0(study_prefix, "_", gsub("/Differential_stats", "", analysis), ifelse(nchar(sub_analysis)>12, paste0("_", sub_analysis), "" ), "_Norm.RData")
  if (RData_normfile %in% dir(RData_results_dir)) {
    load(paste0(RData_results_dir, RData_normfile))
  } else {
    normalise = TRUE
  }
}

```


```{r annotate-illumina-reference-probes}
# **********
# Annotate reference files with SNPs and multimappers defined by McCartney et al 2016, Zhou et al 2017, Nordlund et al 2013
# **********
if (array_type=="EPIC"){
  # **********
  # Get the lists of SNP and multi mapper data for the EPIC array
  # **********
  EPIC.manifest.hg19 = fread(EPIC_hg19_file, na.strings="")
  Zhou_SNP = data.frame(IlmnID = as.character(EPIC.manifest.hg19$probeID[which(EPIC.manifest.hg19$MASK_general==TRUE)]))
  data(multi.hit)
  Nordlunds_multi = data.frame(IlmnID = as.character(multi.hit$TargetID))
  # Nordlunds_multi = as.data.frame(fread(Nordlunds_file))
  McCartney_SNP = fread(SNP_list)
  McCartney_other = fread(rs_ch_list)
  McCartney_multi = fread(Multimapper_list)
}

```
  

```{r set-up-data-with-ChAMP-import-and-filter, include=FALSE}
# **********
# Imports the raw idat files using ChAMP and the initial filter including the detected p-values and the beadcount.
# **********
if (first_run){
  # *************
  # Import and Filter methylation data
  # *************
  import_messages = ""
  meth_import = champ.import(directory=methylation_directory,
                             arraytype=array_type)
  samp_imp_messages = utils::capture.output(champ.import(directory=methylation_directory,
                                                         arraytype=array_type), type="message")
  # capture import filter messages (ie pval, beads).  Keep non-CpG as probes are known signal???????
  import_messages = utils::capture.output(filter.data(import.data=meth_import,
                                                      filter.type="message",
                                                      first.filter=TRUE,
                                                      array.type=array_type),
                                          append=TRUE,
                                          type="message")
  # filter import before normalise ie only probes and samples that fail QC
  meth_filter = filter.data(import.data=meth_import,
                            filter.type="filter",
                            first.filter=TRUE,
                            array.type=array_type)
  meth_import = list(beta = meth_filter$beta,
                     M = meth_filter$M,
                     intensity = meth_filter$intensity,
                     detP = meth_import$detP[which(rownames(meth_import$detP)%in%rownames(meth_filter$beta)),],
                     beadcount = meth_import$beadcount[which(rownames(meth_import$beadcount)%in%rownames(meth_filter$beta)),],
                     Meth = meth_filter$Meth,
                     UnMeth = meth_filter$UnMeth,
                     pd = meth_filter$pd)
  # capture the second filter messages (ie filtering out the non-CpG, SNPs, multi-mappers and XY chromosome)
  filter_messages = ""
  filter_messages = utils::capture.output(filter.data(import.data=meth_import,
                                                      filter.type="message",
                                                      first.filter=FALSE,
                                                      array.type=array_type),
                                          append=TRUE,
                                          type="message")
}

```

\newpage  

# Summary of the SCLC WGS Methylation.

## Importing from idat Files.  
Prior to initial look at the quality of the raw data, it was filtered to remove all probes with a detected p-value > 0.01 and with fewer than 3 beads in at least 5% of samples. Samples were removed if > 10% of their probes failed. 


### Quality Control of Raw Data.  

```{r set-up-sample-info, results='asis'}
# **********
# Set up sample_info with the sentrix id (combined slide and array) 
# **********
new_samp_info = as.data.frame(fread(file=paste0(working_dir, "Sample_Sheets/Fielding_-_Lung_WGS_SCLC_with_MethClust_82SampleSheet_v14.csv")))
rownames(new_samp_info) = new_samp_info$Sample_Name
meth_import$pd = new_samp_info[colnames(meth_import$beta),]
sample_info_levels = set.sample.info.levels(pd= as.data.frame(meth_import$pd),
                                            samp.vars = sample_sheet_cat_all,
                                            variables = sample_categories, 
                                            var.levs = sample_levels)
sample_info = sample_info_levels$sample.info
sample_levels = sample_info_levels$var.levels
# assigns category colours 
colour_cat = list()
colour_cat_border = list()
for (cat in sample_categories){
  colour_cat[[cat]] = c(sample_colours[[cat]][sample_levels[[cat]][which(sample_levels[[cat]]%in%as.character(sample_info[,cat]))]])
  colour_cat_border[[cat]] = c(sample_border_colours[[cat]][sample_levels[[cat]][which(sample_levels[[cat]]%in%as.character(sample_info[,cat]))]])
  sample_levels[[cat]] = sample_levels[[cat]][which(sample_levels[[cat]]%in%as.character(sample_info[,cat]))]
  names(colour_cat_border[[cat]]) = names(colour_cat[[cat]]) = sample_levels[[cat]]
}

```


```{r run-QC-before-normalisation, fig.height=5, fig.width=7, results='asis'}
# **********
# Apply QC on the imported data that has been filtered for possible technically faulty probes and samples 
# **********
QC()

```


```{r save-imported-data}
# **********
# Save the imported data so the above import and filter need not be processed and re-used in future runs
# **********
if (first_run){
  save(meth_import, import_messages, samp_imp_messages, filter_messages, file=paste0(RData_results_dir, study_prefix, "_", gsub("/Differential_stats", "", analysis), ifelse(nchar(sub_analysis)>12, paste0("_", sub_analysis), "" ), "_Raw.RData"))
}

```

\newpage  

## Normalisation and Filtering of Imported Data.  

All probes that are not intended for the analysis are filtered from the data and then the data is normalised using a `r paste(normalisation_methods, collapse=" nomalisation followed by a ") ` normalisation.  

```{r filter-the-data}
# **********
# Filters are applied for probes that are non CpG, SNPs, Multi-mappers and on the XY chromosome.
# **********
meth_filter = filter.data(import.data=meth_import,
                          filter.type="filter",
                          first.filter=FALSE,
                          array.type=array_type)
```


```{r normalise-data-and-report, include=FALSE}
# **********
# Normalisation methods of choice are applied to the beta values for future analyses
# **********
if (normalise) {
  norm_data = meth_filter$beta 
  norm_list = list()
  for (norm_method in normalisation_methods){
    meth_norm = normalise.data(dat=norm_data,
                               rgSet=rgSet,
                               norm.method=norm_method,
                               array.type=array_type,
                               seed=seed)
    if (norm_method == "QN"){
      colnames(meth_norm) = colnames(norm_data)
      rownames(meth_norm) = rownames(norm_data)
    }
    norm_list[[norm_method]] = meth_norm
    norm_data = meth_norm
  }
} else {
  meth_norm = norm_list[[length(norm_list)]]
}

```


```{r create-filtered-normalised-data}
meth_norm = meth_norm[which(rownames(meth_norm) %in% rownames(meth_filter$beta)),]
meth_M_norm = beta2m(meth_norm) 
for (i in 1:nrow(meth_M_norm)){
  meth_M_norm[i,][which(meth_M_norm[i,]==0)]=0.0000001
}

```

\newpage  

### Quality Control of Normalised Data.  

The quality of the data presented is after filtering, normalisation and looks for the possibility of a batch effect associated with the different batches of samples processed at different time points. A singular value decompostition analysis is run to determine if there are any confounding variables or batch effects that require removing.  

```{r run-qc-on-normalised-data, results='asis', fig.width=10, fig.height=8}
# **********
# Run QC on Normalised beta values
# **********
num_norm_meth = 0
for (norm_method in normalisation_methods){
  num_norm_meth = num_norm_meth+1
  QC(raw=FALSE,
     beta=norm_list[[norm_method]],
     Group=sample_info$Sample_Group,
     mdsPlot=FALSE,
     densityPlot=TRUE,
     dendrogram=FALSE,
     PDFplot=FALSE,
     norm.method=num_norm_meth)
  cat(paste("\n","\n  "))
}

```

\newpage  

### Summary after Filtering Data.    
Methylation data imported from idat files resulted in **`r format.number(nrow(meth_import$beta)) `** probes for **`r format.number(ncol(meth_import$beta)) `** samples.  
The data was then filtered on:  
A. Samples with > 10 % failed probes - resulting in `r get.probe.info(message.file=import_messages, message.type="meth_samp", array.type=array_type)$Message `  

B. All probes with:  
  1. A detection p-values > 0.01 - `r get.probe.info(message.file=import_messages, message.type="meth_pvalue", array.type=array_type)$Message `    
  2. Fewer than 3 beads in at least 5% of samples - `r get.probe.info(message.file=import_messages, message.type="meth_bead", array.type=array_type)$Message `    
  3. Non-CpG - `r get.probe.info(message.file=filter_messages, message.type="meth_CpG", array.type=array_type)$Message  `    
  4. X and Y chromosome location - `r get.probe.info(message.file=filter_messages, message.type="meth_XY", array.type=array_type)$Message `    
  5. SNP-relation (i.e. population polymorphisms) based on Zhou et al and McCartney et al - `r get.probe.info(message.file=filter_messages, message.type="meth_SNP", array.type=array_type)$Message `  
  6. Mapping to multiple regions of the genomes based on Zhou et al and McCartney et al - `r get.probe.info(message.file=filter_messages, message.type="meth_multi", array.type=array_type)$Message `  


```{r create-import-filter-summary-table, fig.height = 2.5, fig.width = 7, results='asis' }
# **********
# Sumamrises and tables the filtering process
# *********
summary_filter = data.frame(Filter=c("Detected p-values > 0.01",
                                     "< 3 beads >= 5% of samples per probe",
                                     "All non CpG probes",
                                     "XY chromosome probes",
                                     "SNP-related probes",
                                     "Multi-mapping probes"),
                            Probes_Removed=c(get.probe.info(message.file=import_messages,
                                                           message.type="meth_pvalue",
                                                           array.type=array_type)$Number,
                                            get.probe.info(message.file=import_messages,
                                                           message.type="meth_bead",
                                                           array.type=array_type)$Number,
                                            get.probe.info(message.file=filter_messages,
                                                           message.type="meth_CpG",
                                                           array.type=array_type)$Number,
                                            get.probe.info(message.file=filter_messages,
                                                           message.type="meth_XY",
                                                           array.type=array_type)$Number,
                                            get.probe.info(message.file=filter_messages,
                                                           message.type="meth_SNP",
                                                           array.type=array_type)$Number,
                                            get.probe.info(message.file=filter_messages,
                                                           message.type="meth_multi",
                                                           array.type=array_type)$Number),
                            Total_Removed=rep(NA, 6),
                            stringsAsFactors=FALSE)
summary_filter$Total_Removed = cumsum(summary_filter$Probes_Removed)
for (i in 1:nrow(summary_filter)){
  summary_filter$Total_Removed[i] = format.number(summary_filter$Total_Removed[i])
  summary_filter$Probes_Removed[i] = format.number(summary_filter$Probes_Removed[i])
}
colnames(summary_filter) = gsub("_", " ", colnames(summary_filter))
tt1 = ttheme_default()
tt1$colhead$fg_params$col = "white"
table_col = c("#123456", "#bcd6f1", "#d1e3f5")
tt1$colhead$bg_params$fill = table_col[1]
tt1$core$bg_params$fill = table_col[2:3]
tt1$core$fg_params$fontsize = 10
tt1$colhead$fg_params$fontsize = 10
title = textGrob("Summary of Probes Filtered from Data for the Analysis.", gp=gpar(fontsize=12))
padding = unit(5, "mm")
summary_filter = gtable_add_rows(tableGrob(summary_filter,
                                            theme=tt1, row=NULL),
                                  heights = grobHeight(title) + padding,
                                  pos = 0)
summary_filter = gtable_add_grob(
  summary_filter,
  title,
  1, 1, 1, ncol(summary_filter)
)
grid.arrange(summary_filter)

```

The data for the **`r study_name `** filtered on the above criteria, resulted in **`r format.number(nrow(meth_filter$beta)) `** probes for **`r format.number(ncol(meth_filter$beta)) `** samples.  Samples consist of **`r format.number(length(which(meth_filter$pd$Sample_Group==sample_levels[["Sample_Group"]][1]))) `** `r sample_levels[["Sample_Group"]][1] ` samples and **`r format.number(length(which(meth_filter$pd$Sample_Group==sample_levels[["Sample_Group"]][2]))) `** `r sample_levels[["Sample_Group"]][2] ` samples. 


\newpage  

## Evaluation of Normalised & Filtered Data.  
A singular value decompostition analysis is run to determine if there are any confounding variables or batch effects that require removing.  

```{r evaluate-normalised-data, fig.width=8, fig.height=7, results='asis' }
# **********
# SVD on normalised data to visualise any batch effects and mark them for possible removal.  
# If the variable names stored in batch_var are in possible_covariates, declared in study-options a batch removal is noted
# ********* norm_list$BMIQ
if (length(which(meth_norm==0))>0){
  for (i in 1:nrow(meth_norm)){
  meth_norm[i,][which(meth_norm[i,]==0)]=0.0000001
  }
}
# simplify and order sample information for the paper
col_order = c("Sample_Name", "Sample_Group", "Slide", "Array", "SentrixID", "Batch_Num") 
meth_norm_SVD = SVD(beta=meth_M_norm,
                    pd=sample_info[,col_order],
                    resultsDir=results_dir)
```

\newpage  

### Evaluation of Normalised Filtered Data for Batch Effect.  
If a batch effect was determined then this is removed using R package ComBat and a singular value decompostition analysis is re-run to determine if the confounding variables or batch effects has been removed.  


```{r remove-batch-effect-data-if-there, fig.width=8, fig.height=7, results='asis'}
# **********
# If the variable names stored in batch_var are in possible_covariates, declared in study-options a batch removal is noted
# ********* 
batch_var = names(which(sapply(apply(meth_norm_SVD[1:4,], 2, function(x) which(x<0.001)), function(x) length(x)) > 0))
batch_cor = NULL
for (batch in batch_var){
  if (batch %in% possible_covariates){
    batch_cor = c(batch_cor, batch)
  }
}
if (length(batch_cor) > 0){
  meth_norm_combat = runCombat(beta=meth_norm,
                               pd=sample_info,
                               batchname=batch_cor)
} else {
  meth_norm_combat = NULL
}

```


```{r run-SVD-again-if-batch-correction-required, fig.width=8, fig.height=7, results='asis'}
# **********
# Corrects for any batch effect and re-runs SVD for visual conformation. Convert the finalised normalised beta values to a matrix of M values
# *********
# if batch effects were to be corrected run SVD again to see if they have been removed
if (length(batch_cor) > 0|| !is.null(meth_norm_combat)){
  meth_norm_SVD = SVD(beta=meth_norm_combat, pd=sample_info,  batch=TRUE)
  meth_norm = meth_norm_combat
  # convert normalised beta values into M values
  meth_M_norm = beta2m(meth_norm)
  for (i in 1:nrow(meth_M_norm)){
    meth_M_norm[i,][which(meth_M_norm[i,]==0)]=0.0000001
  }
} else {
  cat("No batch effect was found or removed.  \n\n")
}

```


```{r save-normalisation-data}
# **********
# Save normalisation result to RData file 
# **********
if (normalise){
  save(norm_list, file=paste0(RData_results_dir, study_prefix, "_", gsub("/Differential_stats", "", analysis), ifelse(nchar(sub_analysis)>12, paste0("_", sub_analysis), "" ), "_Norm.RData"))
}

```



\newpage  

# Predictive Immune Signature for `r study_name `.  

```{r use-methylcibersort-prediction-of-cell-composition, fig.show='asis', fig.keep = 'all', fig.width=8, fig.height=8}
# **********
# Code to create the data sets for deconvolution in CIBERSORT
# **********
# library("MethylCIBERSORT")
# data("StromalMatrix_V2")
# data("Legacy_Signatures")
# data_types = study_prefix
# meth_data_mats = list(meth_norm)
# names(meth_data_mats) = data_types
# Sig_types = c("ImmunoCell_Sig")
# RDM_dir = <name of research directory to save files>
# RNASeq_TMM = as.data.frame(fread(file = <TMM normalised RNASEq data>)))
# CIBER_results = paste0(RDM_dir, "CIBERSORT_Results/")
# system(paste0("mkdir -p ", CIBER_results))
# sig_dir = paste0(CIBER_results, "Signatures/")
# system(paste0("mkdir -p ", sig_dir))
# exp_dir = paste0(CIBER_results, "Export_Matrices/")
# system(paste0("mkdir -p ", exp_dir))
# ran different analyses with - which(Stromal_v2.pheno%in%c("CD8", "Treg", "Eos", "Neu", "CD14"))
# immuno_cells = c(1:length(Stromal_v2.pheno)) 
# for (data_type in data_types){
#   meth.data = meth_data_mats[[data_type]]
#   meth.data = meth.data[,which(colnames(meth.data)%in%gsub("_T", "", colnames(RNASeq_TMM)))]
#   for (sig_type in Sig_types){
#     sig_name = paste0(sig_dir, data_type, "_", sig_type, "_allCells_48cohort")
#     exp_name = paste0(exp_dir, data_type, "_", sig_type, "_Export_allCells_48cohort")
#     if (sig_type=="Cancer_Sig"){
#       Int = intersect(rownames(meth.data), Signatures$oesophagus_Signature.txt$NAME)
#       methyl_mat = meth.data[match(Int, rownames(meth.data)),]
#       Sig_mat = Signatures$oesophagus_Signature.txt[match(Int, Signatures$oesophagus_Signature.txt$NAME),]
#       message("     The dimensions of the signature file are ", dim(Sig_mat)[1], " rows and ", dim(Sig_mat)[2], " columns.")
#       write.csv(Sig_mat, paste0(sig_name, "_Signature.txt"), row.names=FALSE)
#     } else if (sig_type=="ImmunoCell_Sig"){
#       Int = intersect(rownames(meth.data), rownames(Stromal_v2))
#       Stromal_v2 = Stromal_v2[, immuno_cells]
#       message("     The dimensions of the Stromal reference are ", dim(Stromal_v2)[1], " rows and ", dim(Stromal_v2)[2], " columns.")
#       Stromal_mat = Stromal_v2[match(Int, rownames(Stromal_v2)),]
#       message("     The dimensions of the signature file are ", dim(Stromal_mat)[1], " rows and ", dim(Stromal_mat)[2], " columns.")
#       methyl_mat = meth.data[match(Int, rownames(meth.data)),]
#       Signature_meth = FeatureSelect.V4(CellLines.matrix = NULL, Heatmap = FALSE, export = TRUE,
#                                         sigName = sig_name,
#                                         Stroma.matrix = Stromal_mat,
#                                         deltaBeta = 0.2, FDR = 0.01, MaxDMRs = 100, Phenotype.stroma = Stromal_v2.pheno[immuno_cells])
#     } else {
#       message("   The ", data_type, " data ", sig_type, " signature was not created.")
#     }
#     if (sig_type == "Cancer_Sig"){
#       probes = Sig_mat$NAME
#     } else {
#       probes = rownames(Signature_meth$SignatureMatrix)
#     }
#     message("     There are ", length(probes), " probes in the export file."    )
#     Prep.CancerType(Beta = methyl_mat, Probes = probes, fname = exp_name)
#   }
# }

```


```{r plot-methylcibersort-prediction-by-cell-composition-by-treatment-survival, fig.show='asis', fig.keep = 'all', fig.width=16, fig.height=12}
# Import MethyCIBERSORT results from file
ciber_type = ""  # all analyses c("", "noEndo", "ImmuneONLY", "Reduced_ImmuneCells", "48cohort")
CIBER_results_dir = paste0(ifelse(nchar(sub_analysis)>12, gsub("Differential_stats/", "", results_dir), results_dir), "Data/CIBERSORTx/CIBERSORTx_Results/")
CIBER_results = as.data.frame(fread(file=paste0(CIBER_results_dir,"CIBERSORTx_Meth_proportions_", ifelse(ciber_type=="", "", paste0(ciber_type, "_")), "Results.csv"))) # noEndo_
CIBER_results %<>% dplyr::filter(!Mixture=="D05_22_020") #
cell_types = colnames(CIBER_results)[(which(colnames(CIBER_results)=="Mixture")+1):(which(colnames(CIBER_results)=="P-value")-1)]
# Define data.frame for plot
cell_type_proportions = NULL 
for (cell_type in cell_types){
  cell_type_data = data.frame(Donor_Label = unique(CIBER_results$Mixture),
                              Cell_Type = rep(cell_type, length(unique(CIBER_results$Mixture))),
                              Proportion = sapply(unique(CIBER_results$Mixture), function(x) CIBER_results[which(CIBER_results$Mixture==x), cell_type]),
                              Survival_Group = unlist(sapply(unique(CIBER_results$Mixture), function(x) sample_info$Survival_Group[which(sample_info$Sample_Name==x)])),
                              Treatment = unlist(sapply(unique(CIBER_results$Mixture), function(x) sample_info$Treatment[which(sample_info$Sample_Name==x)])))
  cell_type_proportions = rbind(cell_type_proportions, cell_type_data)
}
cell_type_proportions$Cell_Type = factor(cell_type_proportions$Cell_Type, levels = cell_types)
cell_type_proportions$Survival_Group = factor(cell_type_proportions$Survival_Group, levels = sample_levels$Survival_Group)
cell_type_proportions$Treatment = factor(cell_type_proportions$Treatment, levels = sample_levels$Treatment)

```



```{r plot-cibersort-prediction-by-cell-composition-treatment-survival, fig.show='asis', fig.keep = 'all', , fig.width=20, fig.height=12}
# Import RNASeq CIBERSORT results from file
batch_apply = TRUE
CIBER_results_dir = paste0(ifelse(nchar(sub_analysis)>12, gsub("Differential_stats/", "", results_dir), results_dir), "Data/CIBERSORTx/CIBERSORTx_Results/")
CIBER_RNASeq_results = as.data.frame(fread(file=paste0(CIBER_results_dir,"CIBERSORTx_RNASeq_proportions_", ifelse(ciber_type=="Reduced_ImmuneCells", paste0(ciber_type, "_"),""), ifelse(batch_apply, "withBatch_",""), "Results.csv")))
CIBER_RNASeq_results %<>% dplyr::filter(!Mixture=="D05_22_020_T")
cell_types = colnames(CIBER_RNASeq_results)
# Define data.frame for plots
cell_rna_proportions_all = NULL
cell_rna_proportions_comb = NULL
donor_list = gsub("_T", "", CIBER_RNASeq_results$Mixture)[which(gsub("_T", "", CIBER_RNASeq_results$Mixture)%in%sample_info$Sample_Name)]
CIBER_RNASeq_results %<>% dplyr::filter(Mixture %in% paste(donor_list, "T", sep="_"))
types_of_interest = unique(cell_types[(which(colnames(CIBER_RNASeq_results)=="Mixture")+1):(which(colnames(CIBER_RNASeq_results)=="P-value")-1)])
for (cell_type in types_of_interest){
  cell_type_data = data.frame(Donor_Label = donor_list,
                              Cell_Type = rep(cell_type, nrow(CIBER_RNASeq_results)),
                              Proportion = sapply(CIBER_RNASeq_results$Mixture, function(x) sum(CIBER_RNASeq_results[which(CIBER_RNASeq_results$Mixture==x), which(colnames(CIBER_RNASeq_results)==cell_type)])), 
                              Survival_Group = unlist(sapply(donor_list, function(x) sample_info$Survival_Group[which(sample_info$Sample_Name==x)])),
                              Treatment = unlist(sapply(donor_list, function(x) sample_info$Treatment[which(sample_info$Sample_Name==x)])))
  cell_rna_proportions_all = rbind(cell_rna_proportions_all, cell_type_data)
}
cell_rna_proportions_all$Cell_Type = factor(cell_rna_proportions_all$Cell_Type, levels = types_of_interest)
cell_rna_proportions_all$Survival_Group = factor(cell_rna_proportions_all$Survival_Group, levels = sample_levels$Survival_Group)
cell_rna_proportions_all$Treatment = factor(cell_rna_proportions_all$Treatment, levels = sample_levels$Treatment)
# code added to only look at those cell types of interest
types_of_interest = unique(gsub(" naive", "", gsub(" memory", "", gsub(" resting" , "", gsub(" activated", "", gsub(" M0", "", gsub(" M1", "", gsub(" M2", "", types_of_interest))))))))
cell_types = gsub(" naive", "", gsub(" memory", "", gsub(" resting" , "", gsub(" activated", "", gsub(" M0", "", gsub(" M1", "", gsub(" M2", "", cell_types)))))))
for (cell_type in types_of_interest){
  cell_type_data = data.frame(Donor_Label = donor_list,
                              Cell_Type = rep(cell_type, nrow(CIBER_RNASeq_results)),
                              Proportion = sapply(CIBER_RNASeq_results$Mixture, function(x) sum(CIBER_RNASeq_results[which(CIBER_RNASeq_results$Mixture==x), which(cell_types==cell_type)])), 
                              Survival_Group = unlist(sapply(donor_list, function(x) sample_info$Survival_Group[which(sample_info$Sample_Name==x)])),
                              Treatment = unlist(sapply(donor_list, function(x) sample_info$Treatment[which(sample_info$Sample_Name==x)])))
  cell_rna_proportions_comb = rbind(cell_rna_proportions_comb, cell_type_data)
}
cell_rna_proportions_comb$Cell_Type = factor(cell_rna_proportions_comb$Cell_Type, levels = types_of_interest)
cell_rna_proportions_comb$Survival_Group = factor(cell_rna_proportions_comb$Survival_Group, levels = sample_levels$Survival_Group)
cell_rna_proportions_comb$Treatment = factor(cell_rna_proportions_comb$Treatment, levels = sample_levels$Treatment)

# Nominate which set to plot
cell_rna_proportions = cell_rna_proportions_all

```

```{r set-up-data-sets-for-diff-analysis-and-cluster-analysis, fig.width=10, fig.height=10, results='asis'}
# **********
# data sets for beta, M and dichotomised
# ****************************
meth_data = list(B_probes = meth_norm, 
                 M_probes = meth_M_norm, 
                 D_probes = t(apply(meth_norm, 1, function(x) ifelse(x>=dich_cutoff,1,0))),
                 samples = sample_info) 

meth_dev_B=NULL
if (dev_type=="mad"){
  meth_dev_B = apply(meth_data$B_probes[,which(sample_info[,"Sample_Group"]%in%c("sclc", "LC_neuro"))], 1, mad)
  meth_dev_M = apply(meth_data$M_probes[,which(sample_info[,"Sample_Group"]%in%c("sclc", "LC_neuro"))], 1, mad)
} else if (dev_type=="sd"){
  meth_dev_B = apply(meth_data$B_probes[,which(sample_info[,"Sample_Group"]%in%c("sclc", "LC_neuro"))], 1, sd)
  meth_dev_M = apply(meth_data$M_probes[,which(sample_info[,"Sample_Group"]%in%c("sclc", "LC_neuro"))], 1, sd)
} else {
  message("To calculate the deviation dev_type needs to be 'sd' for standard deviation or 'mad' for median absolute deviation.")
}
if (!is.null(meth_dev_B)){
  meth_dev_B = meth_dev_B[order(meth_dev_B)]
  # decided to use the probes from the variation of the beta values across all samples
  meth_dev_M = meth_dev_B 
}

```

\newpage  

# Cluster Analysis to Investigate Subtype Classifications.  
A cluster analysis was performed on data to obtain a cluster classification for each sample.  

## Cluster Analysis.  
Prior to clustering one of the following filter steps were applied:  
1. Probes that mapped to methylated regions from Heeke et.al. and a `r ifelse(dev_type=="sd", "standard deviation", "median absolute deviation") ` across the beta values for the Tumour samples with a cut off above `r cutoff `.     
2. The top 2000 probes with a `r ifelse(dev_type=="sd", "standard deviation", "median absolute deviation") ` across the beta values for the Tumour samples with a cut off above `r cutoff `.  
3. Probes that mapped to RNASeq genes of significance from Heeke et.al. and a `r ifelse(dev_type=="sd", "standard deviation", "median absolute deviation") ` across the beta values for the Tumour samples with a cut off above `r cutoff `.     
The dichotomised data for each analysis was based on the following criteria:  
Beta values >=`r dich_cutoff ` were considered methylated and given a value of 1 otherwise they were considered un-methylated and given a value of 0.  

Samples were clustered using Euclidean distance and Ward.D2 method or binary distance measure on the dichotomised beta values. The results of this clustering are represented in dendograms and heatmaps. The heatmaps were clustered by rows only, using either a binary or Euclidean distance measure with Ward's clustering method.  


```{r plot-parameters-and-data}
# set plot parameters
main_cex = 0.9
col_lab_cex = 0.6
branch_width = 3
hmap_col_size = 5
hmap_txt_size = 6.5
# colours for clusters obtained probes in common with RBBS data
samp_clust_col = c("#666666", "#5489cf", "#cf9354", "#44bb9e", "#cf549f", "#cf5c54")
samp_clust_fill = c("#dddddd", "#9bb8df", "#d6b38f", "#9ce2db", "#dea0c6", "#e2a19c")
# colours for most variable probes analysis
samp_var_col = c("#666666", "#43bcb2", "#a63ac5", "#bc434f", "#8cbc43", "#c1a83e")
samp_var_fill = c("#dddddd", "#bbe7e3", "#cf95e0", "#d9959c", "#c6dea2", "#d7c885")

if (sub_analysis==""){
  # define sample group of interest
  samp_gp = "Meth_Cluster"
}

```

\newpage

### Cluster Tumour Samples using Probes in Genomic Regions of Interest.
For this analysis the most variable probes in common with regions using RRBS identified as predicitve of heterogeneity from, Heeke et.al paper were used.  


```{r analyse-sample-clusters-with-k-num-regions, fig.width=12, fig.height=11, fig.align='center', results='asis', fig.keep='last'}
if (sub_analysis==""){
  dat_sets = c("Beta", "M", "Dichot")
  Probe_info %<>% dplyr::filter(!CHR=="")
  chr_levels = c(seq(1,22,1), "X", "Y")
  Probe_info$CHR = factor(Probe_info$CHR, levels=chr_levels)
  Probe_info %<>% arrange(CHR, MAPINFO)
  Probe_info %<>% dplyr::filter(IlmnID%in%rownames(meth_norm))
  # analysis with sd of probes across all samples
  region_probes = NULL
  get_region_probes = TRUE
  if (get_region_probes){
    # ********** analysis with meth regions from paper  **********
    #  ********** the code below reads in the regions from Heeke et.al. that are mapped to GRCh38  **********
    regions_of_interest = as.data.frame(fread(file=paste0(working_dir, "Sample_Sheets/Genes/Heeke_Methylation_Regions.csv")))
    dups = which(duplicated(regions_of_interest$Chr_Range_38))
    for (dup in dups){
      regions_of_interest$Subtype[which(regions_of_interest$Chr_Range_38==regions_of_interest$Chr_Range_38[dup])] = paste(regions_of_interest$Subtype[which(regions_of_interest$Chr_Range_38==regions_of_interest$Chr_Range_38[dup])], collapse=";")
    }
    regions_of_interest = regions_of_interest[which(!duplicated(regions_of_interest$Chr_Range_38)),]
    for (i in 1:nrow(regions_of_interest)){
      # ********** the code below uses the regions from Heeke et.al. that are mapped to GRCh38  **********
      row_probes = dplyr::filter(Probe_info, CHR_hg38==regions_of_interest$CHR[i], (Start_hg38>=regions_of_interest$start_pos[i]|End_hg38>=regions_of_interest$start_pos[i]), (Start_hg38<=regions_of_interest$end_pos[i]|End_hg38<=regions_of_interest$end_pos[i])) #$IlmnID
      if (nrow(row_probes)>0){
        row_probes %<>% dplyr::mutate(Heeke_Chr_Range_38 = rep(regions_of_interest$Chr_Range_38[i], nrow(row_probes)),
                                      Heeke_Subtype = rep(regions_of_interest$Subtype[i], nrow(row_probes))) %>% dplyr::select(Heeke_Subtype, Heeke_Chr_Range_38, all_of(colnames(row_probes)))
        # row_probes %<>% dplyr::select(Heeke_Subtype, Heeke_Chr_Range_38, all(colnames(row_probes)[1:(ncol(row_probes)-2)]))
        region_probes = rbind(region_probes, row_probes)
      }
    }
  }
  # extract probes and methylation information for the probes found in methylation regions and annotate the ones used for clustering.
  if (!is.null(region_probes$IlmnID)){
    r_meth_data = meth_data$B_probes[region_probes$IlmnID,]
    r_clust_probes = apply(r_meth_data, 1, sd)
    region_probe_info = dplyr::mutate(region_probes, Region_Clust_probe = sapply(IlmnID, function(x) ifelse(x%in%names(r_clust_probes[which(r_clust_probes>0.2)]), TRUE, FALSE)))
    r_meth_data = data.frame(IlmnID = rownames(r_meth_data),
                             r_meth_data)
    region_probe_info = merge(region_probe_info, r_meth_data, by="IlmnID")
  }
  # run clustering looking at different number of clusters
  num_clust_probes = 2000
  clust_meth = "ward.D2" # ward.D2, complete, ward.D, average
  samp_clust_meth = list()
  for (dat_set in dat_sets){
    if (dat_set == "Beta"){
      # change cutoff analysis to include the top 1% and 2% most variable promoter probes
      if (is.null(region_probes)){
        meth_clust_dev = tail(meth_dev_B, n=num_clust_probes) # tail(meth_dev_B[which(names(meth_dev_B)%in%biological_probes)], n=num_clust_probes)
        meth_clust_data = meth_data$B_probes[names(meth_clust_dev),]
      } else {
        meth_clust_data = meth_data$B_probes[region_probes$IlmnID,]
        meth_clust_data = meth_clust_data[which(apply(meth_data$B_probes[region_probes$IlmnID,], 1, sd)>0.2),]
      }
      dist_meth = "euclidean"
    } else if (dat_set == "M"){
      if (is.null(region_probes)){
        meth_clust_dev = tail(meth_dev_M, n=num_clust_probes) #tail(meth_dev_M[which(names(meth_dev_M)%in%biological_probes)], n=num_clust_probes)
        meth_clust_data = meth_data$M_probes[names(meth_clust_dev),]
      } else {
        meth_clust_data = meth_data$M_probes[region_probes$IlmnID,]
        meth_clust_data = meth_clust_data[which(apply(meth_data$B_probes[region_probes$IlmnID,], 1, sd)>0.2),]
      }
      dist_meth = "euclidean"
    } else {
      if (is.null(region_probes)){
        b_sd = names(tail(meth_dev_B, n=num_clust_probes)) # names(tail(meth_dev_B[which(names(meth_dev_B)%in%biological_probes)], n=num_clust_probes))
        m_sd = names(tail(meth_dev_M, n=num_clust_probes)) # names(tail(meth_dev_M[which(names(meth_dev_M)%in%biological_probes)], n=num_clust_probes))
        meth_clust_dev = meth_dev_B[union(intersect(b_sd, m_sd), union(b_sd[which(!b_sd%in%m_sd)], m_sd[which(!m_sd%in%b_sd)]))]
        meth_clust_data = meth_data$D_probes[names(meth_clust_dev),]
      } else {
        meth_clust_data = meth_data$D_probes[region_probes$IlmnID,]
        meth_clust_data = meth_clust_data[which(apply(meth_data$B_probes[region_probes$IlmnID,], 1, sd)>0.2),]
      }
      dist_meth = "binary"
    }
    # **********
    # Clusters the tumour samples using binary distance measure and Ward's method for linkage
    # **********
    eval_table = get.dend.fit(logit.mat = meth_clust_data,
                              clust.meth = clust_meth,
                              dist.meth = dist_meth,
                              k.values = c(3,4,5))
    
    grid.arrange(eval_table)
    clust_data = meth_clust_data
    colnames(clust_data) = paste0(colnames(clust_data), "_T")
    samp_clust_meth[[dat_set]] = get.leaf.order (clust_data, 
                                                 dist.method=dist_meth,
                                                 hclust.method=clust_meth,
                                                 k.num=4)
    par(mfrow=c(3,1))
    for (i in 3:5){
      sample.dendogram (dat=meth_clust_data, 
                        samp.type="tumour",
                        probe.type=paste(ifelse(is.null(region_probes), "most variable", "Paper Meth regions"), dat_set, "Values"), #"hypo-methylated biologically relevant"
                        clust.col=samp_clust_col,
                        dist.method = dist_meth,
                        clust.method = clust_meth,
                        k.num = (i),
                        col.lab.cex = col_lab_cex,
                        main.cex = main_cex,
                        branch.wdth = branch_width)
    }
    par(mfrow=c(1,1))
  }
}
```

```{r populate-sample-data-with_clusters-regions, fig.width=9, fig.height=11, fig.align='center', results='asis', fig.keep='last'}
samp_cell_stat = sapply(sample_info$Cellularity, function(x) ifelse(x==0, "0%", ifelse(x<50, "<50%", ifelse(x<75, "50% to <75%", ifelse(x<90, "75% to <90%", ">=90%")))))
names(samp_cell_stat) = sample_info$Sample_Name
new_samp_info = add.samp.cat(cat.name="Tumour_Content",
                             cat.values=samp_cell_stat,
                             cat.levels=c("<50%", "50% to <75%", "75% to <90%", ">=90%"),
                             col=c(colorRamp2(seq(1,5, length=5),rev(brewer.pal(11, "BrBG")[1:5]))(1:4)),
                             col.bord=c(colorRamp2(seq(1,5, length=5),rev(brewer.pal(11, "BrBG")[1:5]))(2:5)),
                             samp.inf=sample_info,
                             samp.cat=sample_categories,
                             samp.type="",
                             samp.lev=sample_levels,
                             cat.cols=colour_cat,
                             cat.col.borders=colour_cat_border,
                             samp.clust = NULL
                             )
new_samp_info$Sample_info$Sample_Name = paste0(new_samp_info$Sample_info$Sample_Name, "_T")
if (sub_analysis==""){
  # best cluster M-values with 4 clusters using euclidean distance and ward.D clustering method
  samp_clust_num=4
  new_samp_info$Sample_Clusters = list()
  for (dat_set in dat_sets){
    samp_clust = samp_clust_meth[[dat_set]]
    new_samp_info$Sample_Clusters[[paste0(ifelse(is.null(region_probes), "Var_ALL_", "Paper_Regions"), ifelse(is.null(region_probes), num_clust_probes, ""), "_", dat_set)]] = samp_clust
    meth_clust = sapply(new_samp_info$Sample_info$Sample_Name, function(x) ifelse(x%in%names(samp_clust), samp_clust[x], "Bgn"))
    new_samp_info = add.samp.cat(cat.name=paste0(ifelse(is.null(region_probes), "Var_ALL_", "Paper_Regions"), ifelse(is.null(region_probes), num_clust_probes, ""), "_", dat_set),
                                 cat.values=meth_clust,
                                 cat.levels=c("Bgn", paste0("Group", c(1:samp_clust_num))),
                                 col=samp_clust_fill[1:(samp_clust_num+1)],
                                 col.bord=samp_clust_col[1:(samp_clust_num+1)],
                                 samp.inf=new_samp_info$Sample_info,
                                 samp.cat=new_samp_info$sample_categories,
                                 samp.type="",
                                 samp.lev=new_samp_info$sample_levels,
                                 cat.cols=new_samp_info$colour_cat,
                                 cat.col.borders=new_samp_info$colour_cat_border,
                                 samp.clust = new_samp_info$Sample_Clusters
    )
  }
}

```


```{r choose-best-fit-from above-result-regions}
# **********
# Plots the heatmap of the probes used in the above block analysis and is semi-supervised in methylation cluster order
# **********
if (sub_analysis==""){
  # best cluster M-values with 4 clusters using euclidean distance and ward.D clustering method
  samp_clust_num=4
  dat_set = "M"
  if (dat_set == "Beta"){
    if (is.null(region_probes)){
      # change cutoff analysis to include the top 1% and 2% most variable promoter probes
      meth_clust_dev = tail(meth_dev_B, n=num_clust_probes) 
      meth_clust_all = meth_data$B_probes[names(meth_clust_dev),]
    } else {
      meth_clust_all = meth_data$B_probes[region_probes$IlmnID,]
      meth_clust_all = meth_clust_all[which(apply(meth_data$B_probes[region_probes$IlmnID,], 1, sd)>0.2),]
    }
    dist_meth = "euclidean"
  } else if (dat_set == "M"){
    if (is.null(region_probes)){
      meth_clust_dev = tail(meth_dev_M, n=num_clust_probes) 
      meth_clust_all = meth_data$M_probes[names(meth_clust_dev),]
    } else {
      meth_clust_all = meth_data$D_probes[region_probes$IlmnID,]
      meth_clust_all = meth_clust_all[which(apply(meth_data$B_probes[region_probes$IlmnID,], 1, sd)>0.2),]
    }
    dist_meth = "euclidean"
  } else {
    if (is.null(region_probes)){
      b_sd = names(tail(meth_dev_B, n=num_clust_probes)) 
      m_sd = names(tail(meth_dev_B, n=num_clust_probes)) 
      meth_clust_dev = meth_dev_B[union(intersect(b_sd, m_sd), union(b_sd[which(!b_sd%in%m_sd)], m_sd[which(!m_sd%in%b_sd)]))]
      meth_clust_all = meth_data$D_probes[names(meth_clust_dev),]
    } else {
      meth_clust_all = meth_data$D_probes[region_probes$IlmnID,]
      meth_clust_all = meth_clust_all[which(apply(meth_data$B_probes[region_probes$IlmnID,], 1, sd)>0.2),]
    }
    dist_meth = "binary"
  }
  colnames(meth_clust_all) = paste0(colnames(meth_clust_all), "_T")
  samp_clust = samp_clust_meth[[dat_set]]
}

```


\newpage

### Heatmaps Supervised by Methylation Clusters on Region Probe Set used for Cluster Analysis.  


```{r plot-sample-heatmaps-regions, fig.width=15, fig.height=13, fig.align='center', results='asis'}

# sort and filter probe information.
sub_categories = c("Meth_Cluster") #, "Tumour_Content", "Treatment", "Survival_Group", "Sample_Group")
if (sub_analysis==""){
  if (is.null(region_probes)){
    clust_probes = Probe_info$IlmnID[which(Probe_info$IlmnID%in%rownames(meth_norm)[which(rownames(meth_norm)%in%names(tail(meth_dev_B, n=num_clust_probes)))])]
  } else {
    clust_probes = rownames(meth_clust_all)
  }
  if (length(grep("benign", sample_info$Sample_Group))>0){
    benign_samp = c(rep("Bgn", length(which(sample_info$Sample_Group=="benign"))))
    names(benign_samp) = sample_info$Sample_Name[which(sample_info$Sample_Group=="benign")]
    samp_clust = c(benign_samp, samp_clust)
  }
  dend_hmaps = sample.dend.heatmap(dat=meth_clust_all[clust_probes,],
                                   meth.norm=meth_norm[clust_probes,],
                                   samp.inf=new_samp_info$Sample_info, #sample_info,NULL
                                   samp.type = "tumour",
                                   samp.clust.col = samp_clust_col,
                                   clust.rows=TRUE,
                                   probe.type = "Paper region",
                                   k.num = samp_clust_num,
                                   hmap.txt = 14, #hmap_txt_size,
                                   hmap.col = hmap_col_size,
                                   samp.cat = new_samp_info$sample_categories[sub_categories],
                                   col.cat = new_samp_info$colour_cat_border,
                                   cutoff = round(min(tail(meth_dev_B, n=num_clust_probes)), 2), 
                                   silent=TRUE,
                                   samp.clust = samp_clust,
                                   hclust.method = clust_meth,
                                   dist.method = dist_meth
                                   )
  grid.arrange(dend_hmaps$beta$gtable)
}

```

\newpage

### Cluster Tumour Samples using Probes in RNASeq Genes of Interest.
For this analysis I used the variables probes located in 156 genes from a set of 181 using RNASeq  identified with machine learning as predictive of heterogeneity from, Heeke et.al paper.  


```{r analyse-sample-clusters-with-k-num-rnaseq, fig.width=12, fig.height=11, fig.align='center', results='asis', fig.keep='last'}
if (sub_analysis==""){
  # **********
  # reading in all genes considered in the analysis
  # **********
  genes_set_data = as.data.frame(fread(paste0(working_dir, "Sample_Sheets/Genes/RNASeq_SCLC_Genes_Of_Interest_flagged_v3.csv")))
  extra_genes = c("CD274", "CTLA4", "PDCD1") # , "HMGB3" on X chromosome won't be in methylation data
  genes_set_data = rbind(genes_set_data,
                         data.frame(Gene_number = seq(700, 699+length(extra_genes), 1),
                                    Hugo_Symbol = extra_genes,
                                    CHR = c("chr9", "chr2", "chr2"),
                                    Status = rep("extra", length(extra_genes)),
                                    RNASeq_TMM = rep(TRUE, length(extra_genes)),
                                    Probe_Info = rep(TRUE, length(extra_genes)))
                         )
  # extract subsets of genes for later investigation
  gene_sets = c("RNASeq", "apm", "fibroblast2", "inflamed", "hla", "sclc", "treatment", "oncogene", "immune", "meth_enzyme", "parpi", "special", "fibroblast", "NE", "non-NE", "PLCG2-high", "other", "cytolytic", "epithelial", "mesenchymal", "extra")
  genes_of_interest = list()
  for (gene_set in gene_sets){
    genes_of_interest[[gene_set]] = dplyr::filter(genes_set_data, grepl(gene_set, Status))
    if (gene_set=="RNASeq"){
      genes_of_interest[[gene_set]] %<>% dplyr::filter(!CHR=="chrX")
      genes_of_interest[[gene_set]] %<>% arrange(as.numeric(gsub("chr", "", CHR)))
    } else if (gene_set == "fibroblast"){
      genes_of_interest[[gene_set]] %<>% dplyr::filter(!grepl("fibroblast2", Status))
    } else if (gene_set == "NE") {
      genes_of_interest[[gene_set]] %<>% dplyr::filter(!grepl("non-NE", Status))
    }
  }
  # get probe information for all genes before filtering for rnaseq genes only
  gene_Probe_info = Probe_info[grep(paste(gsub(";", "|", genes_set_data$Hugo_Symbol), collapse="|"), Probe_info$UCSC_RefGene_Name),]
  gene_Probe_info %<>% mutate(Hugo_Symbol = sapply(UCSC_RefGene_Name, function(x) paste(unique(unlist(strsplit(x, ";")))[grep("LOC", unique(unlist(strsplit(x, ";"))), invert=TRUE)], collapse=";")))
  for (gene_row in 1:nrow(gene_Probe_info)){
    if (length(grep(";", gene_Probe_info$Hugo_Symbol[gene_row]))>0){
      genes = unlist(strsplit(gene_Probe_info$Hugo_Symbol[gene_row], ";"))
      for (gene in genes){
        if (which(genes==gene)==1){
          gene_Probe_info$Hugo_Symbol[gene_row]=gene
        } else {
          gene_Probe_info = rbind(gene_Probe_info, gene_Probe_info[gene_row,])
          gene_Probe_info$Hugo_Symbol[nrow(gene_Probe_info)]=gene
        }
      }
    }
  }
  all_probes=FALSE
  # only use the probes associated with the RNASeq analysis for the methylation clustering.
  gene_probes = rownames(meth_norm)[which(rownames(meth_norm)%in%c(unique(gene_Probe_info$IlmnID[which(gene_Probe_info$Hugo_Symbol%in%genes_of_interest$RNASeq$Hugo_Symbol)])))]
  rna_gene_probe_info = dplyr::filter(gene_Probe_info, IlmnID%in%gene_probes, Hugo_Symbol%in%genes_of_interest$RNASeq$Hugo_Symbol)
  get_gene_number = length(unique(rna_gene_probe_info$Hugo_Symbol[which(rna_gene_probe_info$IlmnID%in%gene_probes)]))
  clust_gene_probes = rownames(meth_norm)[which(rownames(meth_norm)%in%c(unique(rna_gene_probe_info$IlmnID)))]
  if (!all_probes){
    gene_meth_dev_B = apply(meth_data$B_probes[clust_gene_probes,], 1, sd)
    clust_gene_dev = gene_meth_dev_B[which(gene_meth_dev_B>0.2)]
    clust_gene_probes = names(clust_gene_dev)
  } 
  get_clust_gene_number = length(unique(rna_gene_probe_info$Hugo_Symbol[which(rna_gene_probe_info$IlmnID%in%clust_gene_probes)]))
  # get the probe and methylation information used in the analysis
  rna_gene_probe_info = rna_gene_probe_info[which(!duplicated(rna_gene_probe_info$IlmnID)),]
  rna_gene_probe_info = rna_gene_probe_info[, which(!colnames(rna_gene_probe_info)=="Hugo_Symbol")]
  rna_gene_probe_info %<>% dplyr::mutate(RNASeq_Clust_Probe = sapply(IlmnID, function(x) ifelse(x%in%clust_gene_probes, TRUE, FALSE))) %>% dplyr::select(RNASeq_Clust_Probe, all_of(colnames(rna_gene_probe_info)))
  rna_meth_data = meth_norm[which(rownames(meth_norm)%in%gene_probes),]
  rna_meth_data = data.frame(IlmnID = rownames(rna_meth_data),
                             rna_meth_data)
  rna_probe_info = merge(rna_gene_probe_info, rna_meth_data, by="IlmnID")
  gene_clust_meth = list()
  gene_meth_dev_B=NULL
  for (dat_set in dat_sets){
    # gene_probes = rownames(meth_norm)[which(rownames(meth_norm)%in%c(gene_Probe_info$IlmnID))]
    if(dat_set=="M"){
      gene_clust_num = 4
      if (!all_probes){
        gene_probes = clust_gene_probes
      } 
      gene_dat = meth_data$M_probes[gene_probes,]
      dist_meth = "euclidean"
    } else if (dat_set=="Beta") {
      gene_clust_num = 4
      if (!all_probes){
        gene_probes = clust_gene_probes
      } 
      gene_dat = meth_data$B_probes[gene_probes,]
      dist_meth = "euclidean"
    } else {
      gene_clust_num = 4
      if (!all_probes){
        gene_probes = clust_gene_probes
      } 
      gene_dat = meth_data$D_probes[gene_probes,]
      dist_meth = "binary"
    }
    eval_table = get.dend.fit(logit.mat = gene_dat,
                              clust.meth = clust_meth,
                              dist.meth = dist_meth,
                              k.values = c(3,4,5))
    grid.arrange(eval_table)
    clust_data = gene_dat
    colnames(clust_data) = paste0(colnames(clust_data), "_T")
    gene_clust_meth[[dat_set]] = get.leaf.order (clust_data, 
                                                 dist.method=dist_meth,
                                                 hclust.method=clust_meth,
                                                 k.num=gene_clust_num)
    gene_clust_col = c("#666666", "#49b05c", "#5108b4","#f58919", "#d9c80d", "#f188d5", "#610aa3", "#e62843")
    gene_clust_fill = c("#dddddd", "#b9e4c1", "#d3c3ef", "#f8d9b0", "#edeac4", "#fbdff4", "#d2a0f8", "#f49fab")
    # Add methylation clusters to sample information
    gene_clust = gene_clust_meth[[dat_set]]
    new_samp_info$Sample_Clusters[[paste0(nrow(genes_of_interest$RNASeq), "Gene_Meth_", nrow(clust_data), "_", dat_set)]] = gene_clust
    meth_clust = sapply(new_samp_info$Sample_info$Sample_Name, function(x) ifelse(x%in%names(gene_clust), gene_clust[x], "Bgn"))
    new_samp_info = add.samp.cat(cat.name=paste0(nrow(genes_of_interest$RNASeq), "Gene_Meth_", nrow(clust_data), "_", dat_set),
                                 cat.values=meth_clust,
                                 cat.levels=c("Bgn", paste0("Group", c(1:gene_clust_num))),
                                 col=gene_clust_fill[1:(gene_clust_num+1)],
                                 col.bord=gene_clust_col[1:(gene_clust_num+1)],
                                 samp.inf=new_samp_info$Sample_info,
                                 samp.cat=new_samp_info$sample_categories,
                                 samp.type="",
                                 samp.lev=new_samp_info$sample_levels,
                                 cat.cols=new_samp_info$colour_cat,
                                 cat.col.borders=new_samp_info$colour_cat_border,
                                 samp.clust = new_samp_info$Sample_Clusters
                                 )
    par(mfrow=c(3,1))
    for (i in 3:5){
      samp_dend = sample.dendogram (dat=gene_dat,
                                    samp.type="tumour",
                                    probe.type=paste("most variable",  nrow(genes_of_interest$RNASeq), "Gene", dat_set, "Values"), #"hypo-methylated biologically relevant"
                                    clust.col=gene_clust_col[1:(i+1)],
                                    dist.method = dist_meth, 
                                    clust.method = clust_meth,
                                    k.num = (i),
                                    col.lab.cex = col_lab_cex,
                                    main.cex = main_cex,
                                    branch.wdth = branch_width)
    }
    par(mfrow=c(1,1))
  }
  # best cluster M-values with 4 clusters using euclidean distance and ward.D clustering method
  cluster_data = data.frame(Index = c(1:length(gene_clust_meth$M)),
                            Donor_Label = names(gene_clust_meth$M),
                            Meth_Cluster = gene_clust_meth$M)
}

```


\newpage

### Heatmaps Supervised by Methylation Clusters on Probe Set used for Cluster Analysis.  


```{r plot-clusters-sample-heatmaps-rnaseq, fig.width=15, fig.height=13, fig.align='center', results='asis'}
if (sub_analysis==""){
  # **********
  # Plots the heatmap of the probes used in the above block analysis and is semi-supervised in methylation cluster order
  # **********
  # Add methylation clusters to sample information
  gene_clust_num = 4
  gene_dat_set = "M"
  samp_clust = gene_clust_meth[[gene_dat_set]]
  if(gene_dat_set=="M"){
    gene_probes = clust_gene_probes
    gene_dat = meth_data$M_probes[gene_probes,]
    dist_meth = "euclidean"
  } else if (gene_dat_set=="Beta") {
    gene_probes = clust_gene_probes
    gene_dat = meth_data$B_probes[gene_probes,]
    dist_meth = "euclidean"
  } else {
    gene_probes = clust_gene_probes
    gene_dat = meth_data$D_probes[gene_probes,]
    dist_meth = "binary"
  }
  colnames(gene_dat) = paste0(colnames(gene_dat), "_T")
  samp_clust = gene_clust_meth[[gene_dat_set]]
  plot_cat = new_samp_info$sample_categories[c("Tumour_Content", "Treatment", "Survival_Group", "Sample_Group")]
  dend_hmaps = sample.dend.heatmap(dat=gene_dat,
                                   meth.norm=meth_data$B_probes[gene_probes,],
                                   samp.inf=new_samp_info$Sample_info,
                                   samp.type = "tumour",
                                   samp.clust.col = gene_clust_col[1:(gene_clust_num+1)],
                                   clust.rows=TRUE,
                                   probe.levels = gene_probes,
                                   probe.type = ifelse(all_probes, "all relevant", "most variable"),
                                   k.num = (gene_clust_num),
                                   hmap.txt = 14,
                                   hmap.col = hmap_col_size,
                                   samp.cat = plot_cat,
                                   col.cat = new_samp_info$colour_cat_border,
                                   cutoff = paste("Beta >=", round(min(clust_gene_dev), 4)),
                                   new.meth.subroup = "Gene_Meth_Subgroup",
                                   hclust.method = clust_meth,
                                   dist.method = dist_meth,
                                   silent=TRUE,
                                   samp.clust = samp_clust
                                   )
}
genes_in_clust = c()
for (probe in rownames(gene_dat)){
  probe_genes = unique(unlist(strsplit(Probe_info$UCSC_RefGene_Name[which(Probe_info$IlmnID==probe)], ";")))
  probe_genes = probe_genes[which(probe_genes%in%genes_of_interest$RNASeq$Hugo_Symbol)]
  genes_in_clust = unique(c(genes_in_clust, probe_genes))
}
samp_dend = sample.dendogram (dat=gene_dat,
                              samp.type="tumour",
                              probe.type=paste("most variable",  nrow(genes_of_interest$RNASeq), "Gene", dat_set, "Values"), 
                              clust.col=gene_clust_col[1:5],
                              dist.method = dist_meth,
                              clust.method = clust_meth,
                              k.num = (4),
                              col.lab.cex = col_lab_cex,
                              main.cex = main_cex,
                              branch.wdth = branch_width)
# *************************
# Fig 1a
# *************************
grid.arrange(dend_hmaps$beta$gtable)

```


```{r create-gistic-with-clinical }
samp_order = gsub("_T", "", names(samp_clust))
gistic_arm = as.data.frame(fread(file=paste0(working_dir, "Sample_Sheets/GISTIC/Fielding_-_Lung_WGS_GISTIC_by_arm.tsv")))
rownames(gistic_arm) = gistic_arm$`Chromosome Arm`
gistic_arm = gistic_arm[,samp_order[which(samp_order%in%colnames(gistic_arm))]]
loh_arm = as.data.frame(fread(file=paste0(working_dir, "Sample_Sheets/GISTIC/Fielding_-_Lung_WGS_LoH_by_arm.tsv")))
rownames(loh_arm) = loh_arm$Chrom_Arm
loh_arm = loh_arm[,samp_order[which(samp_order%in%colnames(loh_arm))]]
colnames(gistic_arm) = sapply(colnames(gistic_arm), function(x) paste0(x, "_T"))
colnames(loh_arm) = sapply(colnames(loh_arm), function(x) paste0(x, "_T"))
loh_arm = loh_arm[which(rownames(loh_arm)%in%rownames(gistic_arm)),]
samp_df = new_samp_info$Sample_info[which(new_samp_info$Sample_info$Sample_Name%in%colnames(gistic_arm)),]
rownames(samp_df) = samp_df$Sample_Name
samp_df = samp_df[colnames(gistic_arm),]
cn_clin_plots = draw.clinical.annotations(samp.info = samp_df,
                                          samp.cat = c("Sample_Group", "Survival_Group", "Treatment", "Tumour_Content", "Meth_Cluster"),
                                          samp.lev = new_samp_info$sample_levels,
                                          cat.col = new_samp_info$colour_cat_border,
                                          matrix.comparison = "with Copy Number",
                                          hm.matrix = as.matrix(gistic_arm),
                                          hm.matrix.col = colorRamp2(c(-4,0,4), c("blue", "white", "red")),
                                          hm.lab = "Median LogR",
                                          hm.r.pad = 2.5,
                                          extra.matrix = as.matrix(loh_arm),
                                          extra.matrix.col = colorRamp2(c(0,100), c("white", "navy")),
                                          extra.lab = "LoH Percent",
                                          extra.r.pad = 1.5
                                          )

```


```{r plot-hmap-gistic-clinical, fig.height=12, fig.width=13 }
plot_list = list(cn_clin_plots$annot, cn_clin_plots$extra, cn_clin_plots$ht) 
plot_overview = arrangeGrob(grobs = plot_list, layout_matrix = rbind(c(rep(1, 200)), 
                                                                     c(rep(NA, 6), rep(2, 190), rep(NA, 3)), 
                                                                     c(rep(NA, 6), rep(2, 190), rep(NA, 3)), 
                                                                     c(rep(NA, 6), rep(2, 190), rep(NA, 3)), 
                                                                     c(rep(NA, 6), rep(2, 190), rep(NA, 3)), 
                                                                     c(rep(NA, 6), rep(3, 191), rep(NA, 2)), 
                                                                     c(rep(NA, 6), rep(3, 191), rep(NA, 2)), 
                                                                     c(rep(NA, 6), rep(3, 191), rep(NA, 2)), 
                                                                     c(rep(NA, 6), rep(3, 191), rep(NA, 2)) 
                                                                     ))
# *************************
# Fig S7 a & b
# *************************
grid.arrange(plot_overview)

```


```{r examine-chromosomal-arm-loh, fig.show='asis', fig.keep='all', fig.width=10, fig.height=5 }
loh_arms = c("8p", "19p")
loh_arm_p = list()
for (arm in loh_arms){
  plot_data = melt(loh_arm[arm,], value.name = "LOH") %>% dplyr::mutate(Meth_Cluster = sapply(variable, function(x) samp_df$Meth_Cluster[which(samp_df$Sample_Name==x)]))
  gp_levels = unique(samp_df$Meth_Cluster)[order(unique(samp_df$Meth_Cluster))]
  p = ggplot(data=plot_data, aes(x=Meth_Cluster, y=LOH)) + geom_boxplot(aes(fill=Meth_Cluster, color=Meth_Cluster))  + 
    scale_fill_manual(values = new_samp_info$colour_cat$Meth_Cluster) + scale_color_manual(values = new_samp_info$colour_cat_border$Meth_Cluster) + ylim(0,100) +
      labs(title=paste("CHR", arm), y="Percentage LOH", x="Methylation Sub-Group") + 
      theme(panel.background=element_rect(fill='transparent'),
            axis.line = element_line(color="black"),
            axis.text.x.bottom = element_text(size=10), 
            axis.text = element_text(size=12),
            plot.title = element_text(size=14),
            legend.position = "none") +
      stat_compare_means(aes(group=Meth_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), comparisons=NULL) 
  loh_arm_p[[arm]] = ggplotGrob(p)   
}
gt = arrangeGrob(grobs=loh_arm_p, ncol=2)
# *************************
# Fig S7 c & d
# *************************
grid.arrange(gt)

```

```{r plot-methylcibersort-prediction-by-cell-composition-by-cluster, fig.show='asis', fig.keep = 'all', fig.width=25, fig.height=10}
cell_type_proportions %<>% dplyr::mutate(Methylation_Cluster = sapply(Donor_Label, function(x) new_samp_info$Sample_info$Meth_Cluster[which(gsub("_T", "", new_samp_info$Sample_info$Sample_Name)==x)])) 
cell_type_proportions$Methylation_Cluster = factor(cell_type_proportions$Methylation_Cluster, levels = unique(new_samp_info$Sample_info$Meth_Cluster)[order(unique(new_samp_info$Sample_info$Meth_Cluster))])
groups = c("Methylation_Cluster") #, "Survival_Group
gene_clust_col = c(Bng="#666666", `Group1`="#49b05c", `Group2`="#5108b4", `Group3`="#f58919", `Group4`="#d9c80d", `Group5`="#f188d5",`Group6`= "#610aa3", `Group7`="#e62843")
gene_clust_fill = c(Bng="#dddddd", `Group1`="#b9e4c1", `Group2`="#d3c3ef", `Group3`="#f8d9b0", `Group4`="#edeac4", `Group5`="#fbdff4", `Group6`="#d2a0f8", `Group7`="#f49fab")

cell_types = unique(cell_type_proportions$Cell_Type)
meth_cells_p = list()
gp2_cells_p = list()
for (cell_t in cell_types){
  plot_data = dplyr::filter(cell_type_proportions, Cell_Type == cell_t)
  gp_levels = unique(new_samp_info$Sample_info$Meth_Cluster)[order(unique(new_samp_info$Sample_info$Meth_Cluster))]
  p = ggplot(data=plot_data, aes(x=Methylation_Cluster, y=Proportion)) + geom_boxplot(aes(fill=Methylation_Cluster, color=Methylation_Cluster))  + 
    scale_fill_manual(values = gene_clust_fill) + scale_color_manual(values = gene_clust_col) + ylim(0,max(plot_data$Proportion)+0.05*max(plot_data$Proportion)) +
      labs(title=cell_t, y="Proportion Cell Type", x="Methylation Cluster") + 
      theme(panel.background=element_rect(fill='transparent'),
            axis.line = element_line(color="black"),
            axis.text.x.bottom = element_text(size=10), 
            axis.text = element_text(size=12),
            plot.title = element_text(size=14),
            legend.position = "none") +
    stat_compare_means(aes(group=Methylation_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), comparisons = NULL) +
      # stat_compare_means(aes(group=Methylation_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), label="p.format", label.x.npc = "center") +  
      stat_compare_means(aes(group=Methylation_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), label="p.signif", label.x.npc = 0.5)
  meth_cells_p[[cell_t]] = ggplotGrob(p)
  gp2_data = dplyr::filter(plot_data, Methylation_Cluster=="Group2", Survival_Group %in% c("Long", "Short"))
  gp_levels = c("Long", "Short")
  gp2_p = ggplot(data=gp2_data, aes(x=Survival_Group, y=Proportion)) + geom_boxplot(aes(fill=Survival_Group, color=Survival_Group))  +  
    scale_fill_manual(values = new_samp_info$colour_cat$Survival_Group) + scale_color_manual(values = new_samp_info$colour_cat_border$Survival_Group) + ylim(0,max(gp2_data$Proportion)+0.05*max(gp2_data$Proportion)) +
      labs(title=cell_t, y=paste("Proportion of", gsub(" ", "", cell_t), "within Group2"), x="Survival") + 
      theme(panel.background=element_rect(fill='transparent'),
            axis.line = element_line(color="black"),
            axis.text.x.bottom = element_text(size=10), 
            axis.text = element_text(size=12),
            plot.title = element_text(size=14),
            legend.position = "none") +
    stat_compare_means(aes(group=Survival_Group), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), comparisons=NULL) 
  gp2_cells_p[[cell_t]] = ggplotGrob(gp2_p) 
}
gt = arrangeGrob(grobs=meth_cells_p, ncol=5)
# *************************
# Fig 3 a
# *************************
grid.arrange(gt)

```



```{r plot-cibersort-prediction-by-cell-composition-by-cluster, fig.show='asis', fig.keep = 'all', fig.width=40, fig.height=15}
cell_rna_proportions_all %<>% dplyr::mutate(Methylation_Cluster = sapply(Donor_Label, function(x) new_samp_info$Sample_info$Meth_Cluster[which(gsub("_T", "", new_samp_info$Sample_info$Sample_Name)==x)])) 
cell_rna_proportions_all$Methylations_Cluster = factor(cell_rna_proportions_all$Methylation_Cluster, levels = unique(new_samp_info$Sample_info$Meth_Cluster)[order(unique(new_samp_info$Sample_info$Meth_Cluster))])
cell_rna_proportions_comb %<>% dplyr::mutate(Methylation_Cluster = sapply(Donor_Label, function(x) new_samp_info$Sample_info$Meth_Cluster[which(gsub("_T", "", new_samp_info$Sample_info$Sample_Name)==x)])) 
cell_rna_proportions_comb$Methylations_Cluster = factor(cell_rna_proportions_comb$Methylation_Cluster, levels = unique(new_samp_info$Sample_info$Meth_Cluster)[order(unique(new_samp_info$Sample_info$Meth_Cluster))])

```


\newpage  


## Differential Methylation between Sample Groups.    

```{r global-overview-tumour-by-cluster-comparison, fig.width=12, fig.height=16, fig.align='center', fig.keep='last', results='asis' }
# **********
# Differential methylation between the sample groups overview plot
# *********
if (sub_analysis==""){
  group_comparisons = list(T1T2 = c("Group1", "Group2"),
                           T1T3 = c("Group1", "Group3"),
                           T1T4 = c("Group1", "Group4"),
                           T2T3 = c("Group2", "Group3"),
                           T2T4 = c("Group2", "Group4"),
                           T3T4 = c("Group3", "Group4")
                           )
  group = "Meth_Cluster"
} else if (length(grep("longVshort", sub_analysis))>0){
  group = "Survival_Group"
  group_comparisons = list(LongShort = c("Long", "Short"))
  new_samp_info$Sample_info$Survival_Group = factor(new_samp_info$Sample_info$Survival_Group, levels = c("Long", "Short"))
} else if (length(grep("chemIOVother", sub_analysis))>0){
  group = "Treatment"
  group_comparisons = list(LongShort = c("ChemoIO", "Other"))
  new_samp_info$Sample_info$Treatment = factor(sample_info$Treatment, levels = c("chemo/immuno", "other"))
} else {
  group = "Sample_Group"
  group_comparisons = list(LongShort = c("SCLC", "Neuro"))
  new_samp_info$Sample_info$Sample_Group = droplevels(sample_info$Sample_Group)
}
if (file.exists(paste0(Stats_results_dir, study_prefix, "_ChAMP_limma_results.RData"))){
  load(paste0(Stats_results_dir, study_prefix, "_ChAMP_limma_results.RData"))
} else {
  limma_diff_meth = list()
}
sample_groups = new_samp_info$Sample_info
sample_groups$Sample_Name = gsub("_T", "", sample_groups$Sample_Name)
sample_groups$Sample_Name = factor(sample_groups$Sample_Name, levels=colnames(meth_norm))
sample_groups %<>% dplyr::arrange(Sample_Name)
dm_sd = 0.1
delta_beta = 0.2
padj = 0.05
promoter_regions = c("TSS1500", "TSS200", "5'UTR", "1stExon")
summary_stats = NULL
limma_analysis = TRUE
for (gp_comp in names(group_comparisons)){
  dm_norm = meth_data$B_probes[,which(colnames(meth_data$B_probes)%in%sample_groups$Sample_Name[which(sample_groups[,group]%in%group_comparisons[[gp_comp]])])]
  dm_M = meth_data$M_probes[,which(colnames(meth_data$B_probes)%in%sample_groups$Sample_Name[which(sample_groups[,group]%in%group_comparisons[[gp_comp]])])]
  dm_samp_group = sample_groups[which(sample_groups[,group]%in%group_comparisons[[gp_comp]]),]
  if (!is.null(dm_sd)){
    meth_dev_diff_B=NULL
    if (dev_type=="mad"){
      meth_dev_diff_B = apply(dm_norm, 1, mad)
    } else if (dev_type=="sd"){
      meth_dev_diff_B = apply(dm_norm, 1, sd)
    } else {
      message("To calculate the deviation dev_type needs to be 'sd' for standard deviation or 'mad' for median absolute deviation.")
    }
    if (!is.null(meth_dev_diff_B)){
      meth_dev_diff_B = meth_dev_diff_B[which(meth_dev_diff_B>dm_sd)]
      dm_norm = dm_norm[names(meth_dev_diff_B),]
      dm_M = dm_M[names(meth_dev_diff_B),]
    }
  }
  gp_name_A = group_comparisons[[gp_comp]][1]
  gp_name_B = group_comparisons[[gp_comp]][2]
  samplesA = which(dm_samp_group[,group]%in%gp_name_A)
  samplesB = which(dm_samp_group[,group]%in%gp_name_B)
  if (!gp_comp%in%names(limma_diff_meth)){
    limma_diff_meth[[gp_comp]] = champ.DMP(beta=dm_M, pheno=droplevels(dm_samp_group[,group]), adjPVal = padj, arraytype = "EPIC")
    gp_limma = limma_diff_meth[[gp_comp]][[intersect(grep(gp_name_A, names(limma_diff_meth[[gp_comp]])), grep(gp_name_B, names(limma_diff_meth[[gp_comp]])))]]
    gp_limma %<>% dplyr::mutate(Mean_Beta_A = sapply(rownames(gp_limma), function(x) mean(dm_norm[x,samplesA])),
                                Mean_Beta_B = sapply(rownames(gp_limma), function(x) mean(dm_norm[x,samplesB])))
    gp_limma %<>% dplyr::mutate(Log2_FC = log2(Mean_Beta_A/Mean_Beta_B))
    colnames(gp_limma) = gsub("Beta_A", paste0("Beta_", gp_name_A), gsub("Beta_B", paste0("Beta_", gp_name_B), colnames(gp_limma)))
    limma_diff_meth[[gp_comp]][[intersect(grep(gp_name_A, names(limma_diff_meth[[gp_comp]])), grep(gp_name_B, names(limma_diff_meth[[gp_comp]])))]] = gp_limma
  } else {
    gp_limma = limma_diff_meth[[gp_comp]][[intersect(grep(gp_name_A, names(limma_diff_meth[[gp_comp]])), grep(gp_name_B, names(limma_diff_meth[[gp_comp]])))]]
  }
  fullstats_file = paste0(study_prefix, "_", substr(gp_comp,1,2), "_vs_", substr(gp_comp,3,4), "_sd", dm_sd, ifelse(limma_analysis, "_limma_", "_tumour_"), "fullstats.csv")
  stats_file = gsub("full", "new_", fullstats_file)
  if (file.exists(paste0(results_dir, "Differential_stats/Meth_Comparisons_", gp_comp, "/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/", fullstats_file))){
    full_stats = as.data.frame(fread(paste0(results_dir, "Differential_stats/Meth_Comparisons_", gp_comp, "/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/", fullstats_file)))
  } else {
    full_stats = NULL
  }
  if (file.exists(paste0(results_dir, "Differential_stats/Meth_Comparisons_", gp_comp, "/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/", stats_file))){
    new_stats = as.data.frame(fread(paste0(results_dir, "Differential_stats/Meth_Comparisons_", gp_comp, "/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/", stats_file)))
  } else {
    new_stats = NULL
  }
  if (!is.null(full_stats) & !is.null(new_stats)){
    gp_stats = data.frame(Group_comparison = paste0(group_comparisons[[gp_comp]][1], "vs", group_comparisons[[gp_comp]][2]),
                          Num_HypoMeth_Probes = nrow(dplyr::filter(full_stats, Status=="Hypo-methylated")),
                          Num_HyperMeth_Probes = nrow(dplyr::filter(full_stats, Status=="Hyper-methylated")),
                          Total_Probes = length(unique(full_stats$Probe_name)),
                          Num_Hypo_Genes = length(unique(dplyr::filter(new_stats, Status=="Hypo-methylated")$UCSC_RefGene_Name)),
                          Num_Hyper_Genes = length(unique(dplyr::filter(new_stats, Status=="Hyper-methylated")$UCSC_RefGene_Name)),
                          Num_common_genes = length(intersect(unique(dplyr::filter(new_stats, Status=="Hyper-methylated")$UCSC_RefGene_Name), unique(dplyr::filter(new_stats, Status=="Hypo-methylated")$UCSC_RefGene_Name))),
                          Total_Genes = length(unique(new_stats$UCSC_RefGene_Name)),
                          Num_Hypo_Promoter_Genes = length(unique(dplyr::filter(new_stats, Status=="Hypo-methylated", UCSC_RefGene_Group%in%promoter_regions)$UCSC_RefGene_Name)),
                          Num_Hyper_Promoter_Genes = length(unique(dplyr::filter(new_stats, Status=="Hyper-methylated", UCSC_RefGene_Group%in%promoter_regions)$UCSC_RefGene_Name)),
                          Num_common_promoter_genes = length(intersect(unique(dplyr::filter(new_stats, Status=="Hyper-methylated", UCSC_RefGene_Group%in%promoter_regions)$UCSC_RefGene_Name), unique(dplyr::filter(new_stats, Status=="Hypo-methylated", UCSC_RefGene_Group%in%promoter_regions)$UCSC_RefGene_Name))),
                          Total_Promoter_Genes = length(unique(dplyr::filter(new_stats, UCSC_RefGene_Group%in%promoter_regions)$UCSC_RefGene_Name)))
    summary_stats = rbind(summary_stats, gp_stats)
  }
}
if (!is.null(summary_stats)) {
  write_tsv(summary_stats, file=paste0(results_dir, "Differential_stats/", group, "_Comparisons_Summary.tsv"))
}

```


```{r plot-pathways-from-global-overview-meth_clusters-all, fig.width=24, fig.height=16, fig.align='center', fig.keep='last', results='asis' }
# **********
# Differential methylation pathways between the sample groups 
# *********
path_types = c("Hallmark", "C2") # "GO", "Reactome", "Hallmark", "C2" c("Nabet_Other", "Nabet_Hall")
promoter_probes = c("TSS1500", "TSS200", "5'UTR", "1stExon")
seed.value = 553
set.seed(seed.value)
RNASeq_TMM = as.data.frame(fread(paste0(working_dir, "Results/RNA_genes_of_interest_full_48Samp.tsv")))
RNASeq_TMM = RNASeq_TMM[, c(1,2,which(colnames(RNASeq_TMM)%in%new_samp_info$Sample_info$Sample_Name))]
num_mult = 1
all_diff_genes = list()
gene_type = "promoter"
num_diff = list()
num_promoter = list()
num_genes = list()
num_exp = list()
path_fc_delta = list()
# set flag to use mean values, if FALSE will use median values
use_mean = TRUE
# set flag to filter for genes, between the groups, with more than one differentially methylated probe in the promoter region.
probe_filter = FALSE
gene_set_filter = FALSE
for (path_type in path_types){
  for (gp_comp in names(group_comparisons)){
    # import annotated limma stats for genes (all gene regions)
    diff_genes = data.frame(fread(file = paste0(results_dir, "Differential_stats/Meth_Comparisons_", gp_comp, "/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/Fielding_-_Lung_All_Seq_", substring(gp_comp, 1,2), "_vs_", substring(gp_comp, 3,4),
                                                ifelse(!is.null(dm_sd), paste0("_sd", dm_sd),""), ifelse(limma_analysis, "_limma", ""), "_new_stats.csv")))
    num_diff[[paste(path_type, gp_comp, sep="_")]] = length(unique(diff_genes$UCSC_RefGene_Name))
    # filter diff genes to include only probes in the promoter regions
    diff_genes %<>% dplyr::filter(UCSC_RefGene_Group%in%promoter_probes)
    num_promoter[[paste(path_type, gp_comp, sep="_")]] = length(unique(diff_genes$UCSC_RefGene_Name))
    # ********************
    # only retain those genes with a min count across the samples
    # ********************
    gp_samples = new_samp_info$Sample_info$Sample_Name[grep(paste(group_comparisons[[gp_comp]], collapse="|"), new_samp_info$Sample_info$Meth_Cluster)]
    gp_num = c(length(which(new_samp_info$Sample_info$Sample_Name[grep(group_comparisons[[gp_comp]][1], new_samp_info$Sample_info$Meth_Cluster)]%in%colnames(RNASeq_TMM))),
               length(which(new_samp_info$Sample_info$Sample_Name[grep(group_comparisons[[gp_comp]][2], new_samp_info$Sample_info$Meth_Cluster)]%in%colnames(RNASeq_TMM))))
    gp_rna_seq = dplyr::select(RNASeq_TMM, `Associated Gene Name`, all_of(gp_samples[which(gp_samples%in%colnames(RNASeq_TMM))]))
    gp_rna_seq = gp_rna_seq[which(rowSums(gp_rna_seq[, 2:ncol(gp_rna_seq)])>4),] 
    diff_genes %<>% dplyr::filter(UCSC_RefGene_Name%in%gp_rna_seq$`Associated Gene Name`)
    num_exp[[paste(path_type, gp_comp, sep="_")]] = length(unique(diff_genes$UCSC_RefGene_Name))
    delta_values = sapply(unique(diff_genes$UCSC_RefGene_Name), function(x) ifelse(use_mean,
                                                                                   mean((dplyr::filter(diff_genes, UCSC_RefGene_Name==x) %>% dplyr::filter(!duplicated(Probe_name)))$Delta_change),
                                                                                   median((dplyr::filter(diff_genes, UCSC_RefGene_Name==x) %>% dplyr::filter(!duplicated(Probe_name)))$Delta_change)))
    delta_values = delta_values[order(delta_values, decreasing = TRUE)]
    fc_values = NULL
    if ("Log2_FC" %in% colnames(diff_genes)){
      for (ucsc_ref in unique(diff_genes$UCSC_RefGene_Name)){
        if (use_mean){
          gene_values = dplyr::filter(diff_genes, UCSC_RefGene_Name==ucsc_ref)
          fc_values = c(fc_values, mean(gene_values$Log2_FC[which(!duplicated(gene_values$Probe_name))]))
        } else {
          gene_values = dplyr::filter(diff_genes, UCSC_RefGene_Name==ucsc_ref)
          fc_values = c(fc_values, median(gene_values$Log2_FC[which(!duplicated(gene_values$Probe_name))]))
        }
      }
      names(fc_values) = unique(diff_genes$UCSC_RefGene_Name)
    }
    fc_values = fc_values[order(fc_values, decreasing=TRUE)]
    pos_values = seq(0, length(which(fc_values>0))*0.000001, 0.000001)
    neg_values = -1*seq(0, length(which(fc_values<0))*0.000001, 0.000001)
    scale_values = c(pos_values[order(pos_values, decreasing=TRUE)], neg_values)
    scale_values= c(head(scale_values, n=length(which(fc_values>0))), tail(scale_values, n=length(which(fc_values<0))))
    fc_values = (fc_values + scale_values)
    if (!gp_comp%in% names(path_fc_delta)){
      path_fc_delta[[gp_comp]] = data.frame(Hugo_Symbol = names(delta_values),
                                            Delta_value = delta_values,
                                            Log2_FC = sapply(names(delta_values), function(x) fc_values[x]))
    }
    message(paste("### Pathway Analysis of Genes from Differential Analyses in", ifelse(gene_type=="promoter", "Gene Promoter", "All Gene"), "Regions -", substring(gp_comp, 1,2), "vs", substring(gp_comp, 3,4), ".  \n\n", sep=" "))
    num_genes[[paste(path_type, gp_comp, sep="_")]] = length(delta_values)
    perform.pathway.analysis (genes = unique(diff_genes$UCSC_RefGene_Name),
                              delta.values = delta_values,
                              fc.values = fc_values,
                              pathway.type = path_type,
                              subgroup.comp = c(group_comparisons[[gp_comp]][1], group_comparisons[[gp_comp]][2]),
                              gene.type = gene_type,
                              use.mean = use_mean,
                              gene.set.filter = gene_set_filter,
                              min.set = ifelse(path_type=="Hallmark", 30, ifelse(length(grep("Nabet", path_type))>0, 2, 100)), 
                              results.dir = paste0(RData_results_dir, "Plots/Meth_Clust_Diff_Analysis/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/Pathway_Analysis/Promoter_Gene_Regions/", ifelse(use_mean, "Mean", "Median"), "_Values/",
                                                  "No_Filter_", seed.value, "/")
                              )
    # store the edited differentially methylated probe information into one list
    if (!gp_comp%in%names(all_diff_genes)){
      all_diff_genes[[gp_comp]] = diff_genes
    }
  }
}

```



```{r look-at-cluster-methylation-across-pathways-per-cluster, fig.width = 25, fig.height= 30 }
gene_set1 = paste0(working_dir, "Sample_Sheets/Pathway_Gene_Sets/h.all.v2024.1.Hs.symbols.gmt")
hallmark_paths = read.gmt(gene_set1)
gene_set2 = paste0(working_dir, "Sample_Sheets/Pathway_Gene_Sets/c2.all.v2024.1.Hs.symbols.gmt")
c2_paths = read.gmt(gene_set2)
full_paths = rbind(hallmark_paths, c2_paths)
full_paths %<>% dplyr::mutate(Cluster_gene = sapply(gene, function(x) ifelse(x%in%genes_in_clust, "TRUE", "FALSE")),
                              T1T2 = rep("", nrow(full_paths)),
                              T1T3 = rep("", nrow(full_paths)),
                              T1T4 = rep("", nrow(full_paths)),
                              T2T3 = rep("", nrow(full_paths)),
                              T2T4 = rep("", nrow(full_paths)),
                              T3T4 = rep("", nrow(full_paths)))
path_genes = list()
min_path = list()
analysis_of_choice = paste0(ifelse(use_mean, "Mean", "Median"), "_Values/No_Filter_", seed.value , "/")
for (gp in names(group_comparisons)) {
  gp_hall_data = as.data.frame(fread(paste0(RData_results_dir, "Plots/Meth_Clust_Diff_Analysis/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/Pathway_Analysis/Promoter_Gene_Regions/", analysis_of_choice, "Fielding_-_Lung_All_Seq_",
                                            group_comparisons[[gp]][1], "_vs_", group_comparisons[[gp]][2], ifelse(limma_analysis, "_limma", ""), ifelse(use_mean, "_mean", "_median"),
                                            "_no_filter", "_Promoter_Regions_Hallmark_GSEA.csv")))
  gp_c2_data = as.data.frame(fread(paste0(RData_results_dir, "Plots/Meth_Clust_Diff_Analysis/Diff_Analysis_DeltaBeta", delta_beta, "_Padj", padj, "/Pathway_Analysis/Promoter_Gene_Regions/", analysis_of_choice, "Fielding_-_Lung_All_Seq_",
                                          group_comparisons[[gp]][1], "_vs_", group_comparisons[[gp]][2], ifelse(limma_analysis, "_limma", ""), ifelse(use_mean, "_mean", "_median"),
                                          "_no_filter", "_Promoter_Regions_C2_GSEA.csv")))
  gp_data = rbind(gp_hall_data, gp_c2_data)
  for (path in unique(full_paths$term)){
    min_pval = 1
    path_data = dplyr::filter(gp_data, ID==path)
    if (nrow(path_data)>0){
      min_path[[path]] = ifelse(min_pval>path_data$p.adjust, gp, min_path[[path]])
      min_pval = ifelse(min_pval>path_data$p.adjust, path_data$p.adjust, min_pval)
      p_genes = unlist(strsplit(path_data$core_enrichment[which(path_data$ID==path)], "/"))
      if (length(p_genes)>0){
        full_paths[which(full_paths$term==path&full_paths$gene%in%p_genes),gp] = "TRUE"
      }
    }
  }
}
full_paths %<>% dplyr::mutate(T1T2 = sapply(T1T2, function(x) ifelse(x=="TRUE", x, "FALSE")),
                              T1T3 = sapply(T1T3, function(x) ifelse(x=="TRUE", x, "FALSE")),
                              T1T4 = sapply(T1T4, function(x) ifelse(x=="TRUE", x, "FALSE")),
                              T2T3 = sapply(T2T3, function(x) ifelse(x=="TRUE", x, "FALSE")),
                              T2T4 = sapply(T2T4, function(x) ifelse(x=="TRUE", x, "FALSE")),
                              T3T4 = sapply(T3T4, function(x) ifelse(x=="TRUE", x, "FALSE")))
full_paths %<>% dplyr::mutate(Num_comp = apply(full_paths, 1, function(x) length(which(x[grep("T", colnames(full_paths), ignore.case = FALSE)]=="TRUE"))))
get_probes = function(x, gene){
  probes = x$Probe_name[which(x$UCSC_RefGene_Name==gene)]
  return(probes)
}
path_avg = NULL
path_rna_avg = NULL
immune_pways = c("HALLMARK_ALLOGRAFT_REJECTION", "HALLMARK_INTERFERON_GAMMA_RESPONSE", "HALLMARK_INTERFERON_ALPHA_RESPONSE", "ZHENG_BOUND_BY_FOXP3", "MARSON_BOUND_BY_FOXP3_UNSTIMULATED", "MARSON_BOUND_BY_FOXP3_STIMULATED", "REACTOME_ADAPTIVE_IMMUNE_SYSTEM",
                 "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "FULCHER_INFLAMMATORY_RESPONSE_LECTIN_VS_LPS_DN", "MATSUDA_NATURAL_KILLER_DIFFERENTIATION", "REACTOME_CYTOKINE_SIGNALING_IN_IMMUNE_SYSTEM")
immune_pway_names = c("ALLOGRAFT_REJECTION", "INTERFERON_GAMMA_RESPONSE", "INTERFERON_ALPHA_RESPONSE", "FOXP3_TARGETS", "FOXP3_TARGETS_UNSTIMULATED", "FOXP3_TARGETS_STIMULATED", "ADAPTIVE_IMMUNE_SYSTEM",
                 "IL2_STAT5_SIGNALING", "TNFA_SIGNALING_VIA_NFKB", "INFLAMMATORY_RESPONSE", "NATURAL_KILLER_DIFFERENTIATION", "CYTOKINE_SIGNALING_IN_IMMUNE_SYSTEM")
stroma_pways = c("NABA_MATRISOME", "REACTOME_EXTRACELLULAR_MATRIX_ORGANIZATION", "DURAND_STROMA_S_UP")
stroma_pway_names = c("MATRISOME", "EXTRACELLULAR_MATRIX_ORGANIZATION", "STROMA")
tumour_pways = c("BENPORATH_ES_WITH_H3K27ME3", "MEISSNER_BRAIN_HCP_WITH_H3K4ME3_AND_H3K27ME3", "BENPORATH_PRC2_TARGETS","HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", "WONG_ADULT_TISSUE_STEM_MODULE", "NUYTTEN_EZH2_TARGETS_UP",
                 "REACTOME_CELL_CYCLE", "BENPORATH_MYC_MAX_TARGETS", "DANG_BOUND_BY_MYC",  "HALLMARK_MYC_TARGETS_V1", "BENPORATH_NANOG_TARGETS", "BENPORATH_SOX2_TARGETS")
tumour_pway_names = c("H3K27ME3", "H3K4ME3_AND_H3K27ME3", "PRC2_TARGETS", "EMT", "STEM_CELLS", "EZH2_TARGETS",
                      "CELL_CYCLE", "MYC_MAX_TARGETS", "BOUND_BY_MYC", "MYC_TARGETS", "NANOG_TARGETS", "SOX2_TARGETS")
pathways_of_interest = c(immune_pways, stroma_pways, tumour_pways)
pathway_names = c(immune_pway_names, stroma_pway_names, tumour_pway_names)
full_int_paths = dplyr::filter(full_paths, term %in% pathways_of_interest) %>% dplyr::mutate(term = sapply(term, function(x) pathway_names[which(pathways_of_interest==x)]))
RNASeq_TMM = as.data.frame(fread(paste0(working_dir, "Results/RNA_genes_of_interest_full_48Samp.tsv")))
RNASeq_TMM = RNASeq_TMM[, c(1,2,which(colnames(RNASeq_TMM)%in%new_samp_info$Sample_info$Sample_Name))]

M_values = FALSE
for (path in pathway_names){
  path_genes[[path]] = dplyr::filter(full_int_paths, term==path, Num_comp>=1)$gene
  for (gene in path_genes[[path]]){
    # get probes for promoter regions
    gene_probes = unique(unlist(lapply(all_diff_genes, function(x) get_probes(x, gene))))
    if (M_values){
      gene_meth = meth_M_norm[gene_probes,]
    } else {
      gene_meth = meth_norm[gene_probes,]
    }
    if(!is.null(gene_meth)){
      gene_rna = dplyr::filter(RNASeq_TMM, `Associated Gene Name`==gene)
      if (nrow(gene_rna)>1){
        gene_rna = rbind(gene_rna, c(gene_rna$`Gene Name`[1], gene_rna$`Associated Gene Name`[1], apply(gene_rna[, 3:ncol(gene_rna)], 2, max)))
      }
      gene_rna_avg = data.frame(Pathway = path,
                                Gene = gene,
                                Group1 = median(log2(as.numeric(dplyr::select(gene_rna[nrow(gene_rna),], new_samp_info$Sample_info$Sample_Name[intersect(which(new_samp_info$Sample_info$Meth_Cluster=="Group1"), which(new_samp_info$Sample_info$Sample_Name%in%colnames(RNASeq_TMM)))]))+1)), 
                                Group2 = median(log2(as.numeric(dplyr::select(gene_rna[nrow(gene_rna),], new_samp_info$Sample_info$Sample_Name[intersect(which(new_samp_info$Sample_info$Meth_Cluster=="Group2"), which(new_samp_info$Sample_info$Sample_Name%in%colnames(RNASeq_TMM)))]))+1)),
                                Group3 = median(log2(as.numeric(dplyr::select(gene_rna[nrow(gene_rna),], new_samp_info$Sample_info$Sample_Name[intersect(which(new_samp_info$Sample_info$Meth_Cluster=="Group3"), which(new_samp_info$Sample_info$Sample_Name%in%colnames(RNASeq_TMM)))]))+1)),
                                Group4 = median(log2(as.numeric(dplyr::select(gene_rna[nrow(gene_rna),], new_samp_info$Sample_info$Sample_Name[intersect(which(new_samp_info$Sample_info$Meth_Cluster=="Group4"), which(new_samp_info$Sample_info$Sample_Name%in%colnames(RNASeq_TMM)))]))+1)))
      path_rna_avg = rbind(path_rna_avg, gene_rna_avg)
      if (is.matrix(gene_meth)){
        if (nrow(gene_meth)>1){
          gene_avg = data.frame(Pathway = path,
                                Gene = gene,
                                Group1 = median(apply(gene_meth[ ,gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group1")])], 1, median)),
                                Group2 = median(apply(gene_meth[ ,gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group2")])], 1, median)),
                                Group3 = median(apply(gene_meth[ ,gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group3")])], 1, median)),
                                Group4 = median(apply(gene_meth[ ,gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group4")])], 1, median)))
          path_avg = rbind(path_avg, gene_avg)
        }
      } else if (is.vector(gene_meth)) {
        gene_avg = data.frame(Pathway = path,
                              Gene = gene,
                              Group1 = median(gene_meth[gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group1")])]),
                              Group2 = median(gene_meth[gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group2")])]),
                              Group3 = median(gene_meth[gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group3")])]),
                              Group4 = median(gene_meth[gsub("_T", "", new_samp_info$Sample_info$Sample_Name[which(new_samp_info$Sample_info$Meth_Cluster=="Group4")])]))
        path_avg = rbind(path_avg, gene_avg)
      }
    }
  }
}
path_types = unique(path_avg$Pathway)
path_rna_avg %<>% dplyr::filter(!is.na(Group1))

```


```{r plot-cluster-methylation-across-pathways-per-cluster-boxplots, fig.width = 25, fig.height= 30 }
sep_path = list()
rna_path = list()
meth_paths_df = NULL
rna_paths_df = NULL
for (path_t in path_types){
  h_path_num = nrow(dplyr::filter(full_int_paths, term==path_t))
  p_path_num = nrow(dplyr::filter(path_avg, Pathway==path_t))
  r_path_num = nrow(dplyr::filter(path_rna_avg, Pathway==path_t))
  plot_data = melt(dplyr::filter(path_avg, Pathway==path_t), variable.name = "Meth_Cluster", value.name="Median_Beta")
  meth_paths_df = rbind(meth_paths_df,
                        data.frame(Pathway = unique(plot_data$Pathway),
                                   Group1_Path_Med = median(plot_data$Median_Beta[which(plot_data$Meth_Cluster=="Group1")]),
                                   Group2_Path_Med = median(plot_data$Median_Beta[which(plot_data$Meth_Cluster=="Group2")]),
                                   Group3_Path_Med = median(plot_data$Median_Beta[which(plot_data$Meth_Cluster=="Group3")]),
                                   Group4_Path_Med = median(plot_data$Median_Beta[which(plot_data$Meth_Cluster=="Group4")])))
  plot_rna_data = melt(dplyr::filter(path_rna_avg, Pathway==path_t), variable.name = "Meth_Cluster", value.name="Median_Log2+1")
  rna_paths_df = rbind(rna_paths_df,
                       data.frame(Pathway = unique(plot_rna_data$Pathway),
                                  Group1_Path_Med = median(plot_rna_data$`Median_Log2+1`[which(plot_rna_data$Meth_Cluster=="Group1")]),
                                  Group2_Path_Med = median(plot_rna_data$`Median_Log2+1`[which(plot_rna_data$Meth_Cluster=="Group2")]),
                                  Group3_Path_Med = median(plot_rna_data$`Median_Log2+1`[which(plot_rna_data$Meth_Cluster=="Group3")]),
                                  Group4_Path_Med = median(plot_rna_data$`Median_Log2+1`[which(plot_rna_data$Meth_Cluster=="Group4")])))
}

```

```{r pathways-heatmap, fig.height=14, fig.width=15}
path_hmaps = list()
path_mat = list()
for (p_hmap in c("Meth", "RNA")){
  if (p_hmap == "Meth") {
    # create zscores for the methylation data
    path_mat[[p_hmap]] = as.matrix(meth_paths_df[2:ncol(meth_paths_df)])
    rownames(path_mat[[p_hmap]]) = gsub("_", " ", meth_paths_df$Pathway)
  } else {
    # create zscores for the RNA data
    path_mat[[p_hmap]] = as.matrix(rna_paths_df[2:ncol(rna_paths_df)])
    rownames(path_mat[[p_hmap]]) = gsub("_", " ", rna_paths_df$Pathway)
  }
  colnames(path_mat[[p_hmap]]) = gsub("_Path_Med", "", colnames(path_mat[[p_hmap]]))
  path_mat[[p_hmap]] = t(apply(path_mat[[p_hmap]], 1, function(x) (x-mean(x))/sd(x)))
  path_col = c("lightblue1", "cadetblue3", "black", "tan4", "darkgoldenrod2")
  path_hmaps[[p_hmap]] = grid.grabExpr(draw(Heatmap(path_mat[[p_hmap]],
                                                    col=colorRamp2(c(-2, -1, 0, 1, 2), path_col),
                                                    cluster_rows=FALSE,
                                                    cluster_columns=FALSE,
                                                    column_title_gp = gpar(fontsize=18),
                                                    row_names_side="left",
                                                    row_title_gp = gpar(fontsize=24),
                                                    row_names_gp=gpar(fontsize=20),
                                                    row_names_max_width = unit(15, "cm"),
                                                    row_split = c(rep("Immune", length(immune_pways)), rep("Stroma", length(stroma_pways)), rep("Tumour", length(tumour_pways))),
                                                    row_gap = unit(10, "mm"),
                                                    column_title=paste(gsub(" DNA Methylation", "", study_name), "-", ifelse(p_hmap=="RNA", "RNA", "Methylation"), "Pathway Scores"),
                                                    column_names_centered = TRUE,
                                                    column_names_gp = gpar(fontsize=20),
                                                    column_names_rot = 0,
                                                    show_column_names = TRUE,
                                                    heatmap_legend_param = list(title="Z-score",
                                                                                title_gp=gpar(fontsize=18, fontface="bold"),
                                                                                labels_gp=gpar(fontsize=14),
                                                                                legend_height=unit(4,"cm"),
                                                                                grid_width=unit(8,"mm"),
                                                                                just=c("right", "top"))), 
                                            padding = unit(c(1,18,1,10),"mm")))
}
# *************************
# Fig 2 b
# *************************
grid.arrange(path_hmaps$Meth)
grid.arrange(path_hmaps$RNA)

```



```{r create-heatmap-annotations-of-all-clusters, fig.width=20, fig.height=4}
col.annotations = dplyr::select(dplyr::arrange(new_samp_info$Sample_info, Meth_Index), Treatment, Survival_Group, Tumour_Content, Sample_Group, Paper_Regions_M, Meth_Cluster) # Var_ALL_2000_M
rownames(col.annotations) = new_samp_info$Sample_info$Sample_Name
hm.zero = matrix(nrow=0, ncol = nrow(col.annotations))
colnames(hm.zero) = rownames(col.annotations)
anno.cols = list()
left.pad = 15
right.pad = 5
for (col in colnames(col.annotations)){
  anno.cols[[col]] = new_samp_info$colour_cat_border[[col]]
}

ha = HeatmapAnnotation(df = col.annotations,
                       col = anno.cols,
                       annotation_height=unit(10, "mm"),
                       annotation_width=unit(10, "mm"),
                       annotation_name_gp = gpar(cex=0.9),
                       show_legend = FALSE
                       )
clust_hm = Heatmap(hm.zero,
                   name = "All Sample Clusters",
                   cluster_rows=FALSE,
                   cluster_columns=FALSE,
                   row_names_side="left",
                   row_names_gp=gpar(fontsize=6),
                   bottom_annotation=ha,
                   show_column_names = FALSE, #ifelse(length(samp.cat)<=16, TRUE, FALSE),
                   # column_names_gp = gpar(fontsize=samp.font.size),
                   column_split = factor(samp_clust),
                   show_heatmap_legend = FALSE
                   )
clust_grob = grid.grabExpr(draw(clust_hm, padding = unit(c(1,left.pad,1,right.pad),"mm")))
# *************************
# Fig S2
# *************************
plot_grid(clust_grob)

```
  

\newpage  
 
# Examine Methylation of Specific Genes of Interest.  


```{r set-up-gene-data, fig.width=8, fig.height=4, fig.align='center' }
# **********
# for methylation data
# **********
genes_for_gene_info = rbind(genes_of_interest$sclc, genes_of_interest$treatment, genes_of_interest$oncogene, genes_of_interest$immune, genes_of_interest$meth_enzyme, genes_of_interest$parpi, genes_of_interest$special, genes_of_interest$fibroblast2, 
                            genes_of_interest$amp, dplyr::filter(genes_set_data, Hugo_Symbol=="REST"), genes_of_interest$cytolytic, genes_of_interest$apm, genes_of_interest$epithelial, genes_of_interest$mesenchymal, genes_of_interest$extra)
genes_for_gene_info = genes_for_gene_info[!duplicated(genes_for_gene_info),]
if (!is.null(genes_for_gene_info)){
  gene_info = list()
  for (gene in genes_for_gene_info$Hugo_Symbol){
    chr_probe_info = dplyr::filter(Probe_info,
                                   IlmnID %in% gene_Probe_info$IlmnID[which(gene_Probe_info$Hugo_Symbol==gene)])
    if (length(grep(gene, as.character(chr_probe_info$UCSC_RefGene_Name)))>0){
      message(paste("Loading gene information for:", gene))
      gene_info[[gene]] = load.gene.info(chr_probe_info, gene)
    }
  }
}
genes_filtered_1 = filter(genes_for_gene_info, !Hugo_Symbol%in%names(gene_info)) %>%
  mutate(Reason = "No entry in probe information file")

genes_for_gene_info %<>% filter(Hugo_Symbol%in%names(gene_info))

```

\newpage

## Heatmaps Specific Genes of Interest.

In this section heatmaps for the list of genes of interest are created.  The heatmaps are supervised, based on probes ordered by chromosome position and the samples ordered on their tumour copy number classification.  

```{r heatmap-of-gene-probes-by-gene-gp, fig.width=16, fig.height=12, fig.align='center', results='asis' }
# check methylation data has suffix in cluster samples
if(length(grep("_T", colnames(meth_norm)))==0){
  colnames(meth_norm) = paste0(colnames(meth_norm), "_T")
}
RNASeq_TMM = as.data.frame(fread(paste0(working_dir, "Results/RNA_genes_of_interest_full_48Samp.tsv")))
RNASeq_TMM_all = as.data.frame(fread(paste0(working_dir, "Results/RNA_genes_of_interest_full_111Samp.tsv")))
# two transcripts for TAP2 removing transcript with fewer than 5 counts across all SCLC samples
RNASeq_TMM_all = RNASeq_TMM_all[which(!RNASeq_TMM_all$`Gene Name` == "ENSG00000250264"),]
rna_study_data = as.data.frame(fread(paste0(working_dir, "Results/RNA_AllLung_samples_withRNA.tsv")))
rna_study_data = rna_study_data[which(!duplicated(rna_study_data)),]
# remove 2 samples that are duplicated so all donors are unique
rna_study_data %<>% dplyr::filter(!Sample%in%c("4400119", "4400067"))
if(length(grep("_T", colnames(RNASeq_TMM_all)))==0){
  colnames(RNASeq_TMM_all)[3:ncol(RNASeq_TMM_all)] = paste0(colnames(RNASeq_TMM_all)[3:ncol(RNASeq_TMM_all)], "_T")
}
if(length(grep("_T", colnames(RNASeq_TMM)))==0){
  colnames(RNASeq_TMM)[3:ncol(RNASeq_TMM)] = paste0(colnames(RNASeq_TMM)[3:ncol(RNASeq_TMM)], "_T")
}
gene_samp_info = new_samp_info$Sample_info
rna_cluster_data = dplyr::filter(cluster_data, Donor_Label%in%colnames(RNASeq_TMM))
rna_gene_samp_info = gene_samp_info[rna_cluster_data$Donor_Label, ]
rna_gene_samp_info$Sample_Name = factor(rna_gene_samp_info$Sample_Name, levels = rna_gene_samp_info$Sample_Name)
rna_meth_norm = meth_norm[,rna_cluster_data$Donor_Label]
RNASeq_TMM = RNASeq_TMM_all[, c(colnames(RNASeq_TMM_all)[c(1,2)],colnames(rna_meth_norm))]
gene_gp_heatmap = list()
clust.name = "Meth_Cluster"
hm_gene_sets = c("sclc")
gene_samp_info = gene_samp_info[names(new_samp_info$Sample_Clusters[[clust.name]]),]
gene_samp_info$Sample_Name = factor(gene_samp_info$Sample_Name, levels = names(new_samp_info$Sample_Clusters[[clust.name]]))
z_score = FALSE
add_annos = c("Survival_Group", "Tumour_Content") 
for (gene_set in hm_gene_sets){
  gene_set_genes = dplyr::filter(genes_of_interest[[gene_set]], !CHR=="chrX") %>% dplyr::arrange(as.numeric(gsub("chr", "", CHR)))
  if (gene_set == "sclc") {
    gene_set_genes %<>% dplyr::filter(!Hugo_Symbol=="REST")
  }
  gene_gp_heatmap[[gene_set]] = draw.gene.heatmap(meth.data = rna_meth_norm,
                                                  samp.info = rna_gene_samp_info,
                                                  RNA.info = RNASeq_TMM,
                                                  cat.cols = new_samp_info$colour_cat_border,
                                                  gene.info = gene_info[c(gene_set_genes$Hugo_Symbol)],
                                                  gene = ifelse(length(gene_set_genes$Hugo_Symbol)>1, "all", gene_set_genes$Hugo_Symbol),
                                                  meth.sub = clust.name,
                                                  gene.meth.sub = ifelse(!is.null(region_probes), "Paper_Regions_M", "Var_ALL_2000_M"), 
                                                  status = gene_set,
                                                  zscore = z_score,
                                                  add.annos = add_annos)
}
# *************************
# Fig 2 a
# *************************
grid.arrange(gene_gp_heatmap$sclc$heatmap)

```


```{r create-rna-seq-heatmaps }
hm_plots = c("Meth_Cluster")
rna_gene_sets = c("NE", "non-NE")
z_score=TRUE
meth_var_order = FALSE
if (meth_var_order){
  meth_var = ifelse(!is.null(region_probes), "Paper_Regions_M", "Var_ALL_2000_M")
  new_samp_info$Sample_info$Sample_Name = factor(new_samp_info$Sample_info$Sample_Name, levels=names(new_samp_info$Sample_Clusters[[meth_var]]))
  new_samp_info$Sample_info %<>% dplyr::arrange(Sample_Name)
  annotations = data.frame(Histology = new_samp_info$Sample_info$Sample_Group,
                           Sub_Group = new_samp_info$Sample_info[,meth_var],
                           Meth_Cluster = new_samp_info$Sample_info$Meth_Cluster
                           )
  rownames(annotations) = new_samp_info$Sample_info$Sample_Name
  anno.cols = new_samp_info$colour_cat_border[c("Meth_Cluster", "Sample_Group", meth_var)]
  names(anno.cols) = c("Meth_Cluster", "Histology", "Sub_Group")
} else {
  meth_var = "Meth_Cluster"
  annotations = data.frame(Survival = new_samp_info$Sample_info$Survival_Group[order(new_samp_info$Sample_info$Meth_Index)],
                           Histology = new_samp_info$Sample_info$Sample_Group[order(new_samp_info$Sample_info$Meth_Index)],
                           Sub_Group = new_samp_info$Sample_info$Meth_Cluster[order(new_samp_info$Sample_info$Meth_Index)]
                           )
  rownames(annotations) = new_samp_info$Sample_info$Sample_Name[order(new_samp_info$Sample_info$Meth_Index)]
  anno.cols = new_samp_info$colour_cat_border[c("Meth_Cluster", "Sample_Group", "Survival_Group")]
  names(anno.cols) = c("Sub_Group", "Histology", "Survival")
}
annotations = annotations[which(rownames(annotations)%in%colnames(RNASeq_TMM)),]
ha = HeatmapAnnotation(df = annotations,
                       col=anno.cols,
                       show_legend=TRUE,
                       show_annotation_name = TRUE,
                       annotation_name_side = "left",
                       annotation_height=unit(ncol(annotations)*4, "mm"),
                       which = "column",
                       annotation_name_gp = gpar(fontsize=10),
                       annotation_legend_param = list(labels_gp = gpar(fontsize=8),
                                                      title_gp = gpar(fontsize=10, fontface="bold"),
                                                      grid_height=unit(5,"mm"),
                                                      grid_width=unit(5,"mm")),
                       na_col = "#dddddd")
rna_hmaps = list()
rna_seq_data = list()
for (gene_set in rna_gene_sets){
  rna_seq_data[[gene_set]] = dplyr::filter(RNASeq_TMM, `Associated Gene Name`%in%c(genes_of_interest[[gene_set]]$Hugo_Symbol))
  plot_rna_seq = as.data.frame(t(rna_seq_data[[gene_set]][,3:ncol(rna_seq_data[[gene_set]])]))
  colnames(plot_rna_seq) = rna_seq_data[[gene_set]]$`Associated Gene Name`
  if (z_score){
    plot_rna_seq = as.data.frame(apply(plot_rna_seq, 2, function(x) (x-mean(x))/sd(x)))
  } else {
    plot_rna_seq = as.data.frame(apply(plot_rna_seq, 2, function(x) log2(x+1)))
  }
  plot_rna_seq %<>% dplyr::mutate(Sample_Name = rownames(plot_rna_seq))
  plot_data = plot_rna_seq
  for (gene in rna_seq_data[[gene_set]]$`Associated Gene Name`){
    plot_data[which(is.na(plot_data[,gene])),gene] = 0
  }
  sample_names = plot_data$Sample_Name
  plot_data = t(dplyr::select(plot_data, colnames(plot_data)[which(colnames(plot_data)%in%c(genes_of_interest[[gene_set]]$Hugo_Symbol))]))
  colnames(plot_data) = sample_names
  hm_plots = "Meth_Cluster"
  for (hm_plot in hm_plots) {
    hm_hmaps = list()
    hm_plot_data = plot_data[, rownames(annotations)]
    hm_plot_data = hm_plot_data[, which(colnames(hm_plot_data)%in%colnames(rna_seq_data[[gene_set]]))]
    if (z_score){
      hm_colours = c("dodgerblue4", "white", "firebrick") 
    } else {
      hm_colours = c("navy", "darkcyan", "gold")
    }
    if (gene_set=="NE"){
      hm_hmaps[[hm_plot]] = grid.grabExpr(draw(Heatmap(hm_plot_data[, which(colnames(hm_plot_data)%in%colnames(rna_seq_data[[gene_set]]))],
                                                       col=colorRamp2(c(ifelse(z_score, -2, 0), ifelse(z_score, 0, 5), ifelse(z_score, 4, 9)), hm_colours),
                                                       cluster_rows=FALSE, 
                                                       cluster_columns=FALSE,
                                                       row_names_side="left",
                                                       row_names_gp=gpar(fontsize=8),
                                                       column_split = new_samp_info$Sample_Clusters[[hm_plot]][which(names(new_samp_info$Sample_Clusters[[hm_plot]])%in%colnames(rna_seq_data[[gene_set]]))],
                                                       column_title=paste(study_name, "-", "RNASeq for", toupper(gene_set), "genes by", gsub("_", "", hm_plot)),
                                                       column_title_gp = gpar(fontsize=10),
                                                       show_column_names = FALSE,
                                                       top_annotation = ha,
                                                       heatmap_legend_param = list(title=ifelse(z_score, "RNA Z-score", "Log2(exp+1)"),
                                                                                   title_gp=gpar(fontsize=10, fontface="bold"), 
                                                                                   labels_gp=gpar(fontsize=8), 
                                                                                   legend_height=unit(5,"cm"), 
                                                                                   grid_width=unit(5,"mm"),
                                                                                   just=c("right", "top")))))
    } else {
      hm_hmaps[[hm_plot]] = grid.grabExpr(draw(Heatmap(hm_plot_data[, which(colnames(hm_plot_data)%in%colnames(rna_seq_data[[gene_set]]))],
                                                       col=colorRamp2(c(ifelse(z_score, -2, 0), ifelse(z_score, 0, 5), ifelse(z_score, 4, 9)), hm_colours),
                                                       cluster_rows=FALSE, 
                                                       cluster_columns=FALSE,
                                                       row_names_side="left",
                                                       row_names_gp=gpar(fontsize=8),
                                                       column_split = new_samp_info$Sample_Clusters[[hm_plot]][which(names(new_samp_info$Sample_Clusters[[hm_plot]])%in%colnames(rna_seq_data[[gene_set]]))],
                                                       column_title=paste(study_name, "-", "RNASeq for", toupper(gene_set), "genes by", gsub("_", "", hm_plot)),
                                                       column_title_gp = gpar(fontsize=10),
                                                       show_column_names = FALSE,
                                                       heatmap_legend_param = list(title=ifelse(z_score, "RNA Z-score", "Log2(exp+1)"),
                                                                                   title_gp=gpar(fontsize=10, fontface="bold"), 
                                                                                   labels_gp=gpar(fontsize=8), 
                                                                                   legend_height=unit(5,"cm"), 
                                                                                   grid_width=unit(5,"mm"),
                                                                                   just=c("right", "top")))))
    }
  }
  rna_hmaps[[gene_set]] = hm_hmaps
}

```



```{r plo-rna-seq-gene-gps, fig.height=12, fig.width=13 }
for (hm_plot in hm_plots) {
  plot1.rh = 4
  plot1.cent = 183
  plot1.lh = 13
  plot_list = list(rna_hmaps$NE[[hm_plot]], rna_hmaps$`non-NE`[[hm_plot]]) 
  plot_overview = arrangeGrob(grobs = plot_list, layout_matrix = rbind(c(rep(NA, 1), rep(1, 199)),  
                                                                       c(rep(NA, 1), rep(1, 199)), 
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, 1), rep(1, 199)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)),
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh)), 
                                                                       c(rep(NA, plot1.rh), rep(2, plot1.cent), rep(NA, plot1.lh))
                                                                       ))
  cat(paste("### Gene Methylation with RNASEQ by Cluster analysis - ", hm_plot, "\n\n", sep=" "))
  # *************************
  # Fig 4 a
  # *************************
  grid.arrange(plot_overview)
  cat(paste("\n\n", "\\pagebreak  ", "\n\n"))
}
```

```{r create-mean-exp-for-nonNE-and-NE-gene-sets, fig.width=12, fig.height=6 }
gene_set_rna_med = NULL
for (gene_set in names(rna_seq_data)) {
  plot_data = data.frame(Sample_Name = colnames(rna_seq_data[[gene_set]])[3:ncol(rna_seq_data[[gene_set]])],
                         RNA_value = sapply(colnames(rna_seq_data[[gene_set]])[3:ncol(rna_seq_data[[gene_set]])], function(x) mean(log2(rna_seq_data[[gene_set]][,x]+1))))
  colnames(plot_data)[which(colnames(plot_data)=="RNA_value")] = gene_set
  if (!is.null(gene_set_rna_med)){
    gene_set_rna_med = merge(gene_set_rna_med, plot_data, by="Sample_Name")
  } else {
    gene_set_rna_med = plot_data
  }
}
gene_set_rna_med %<>% dplyr::mutate(Meth_Cluster = sapply(gene_set_rna_med$Sample_Name, function(x) new_samp_info$Sample_info[,meth_var][which(new_samp_info$Sample_info$Sample_Name==x)]))
gene_set_comp_mean = list()
for (gene_set in names(rna_seq_data)){
  gene_set_rna_med %<>% dplyr::mutate(Gene_Set = gene_set_rna_med[, gene_set])
  gp_levels = unique(new_samp_info$Sample_info$Meth_Cluster)[order(unique(new_samp_info$Sample_info$Meth_Cluster))]
  p = ggplot(data=gene_set_rna_med, aes(x=Meth_Cluster, y=Gene_Set)) + geom_boxplot(aes(fill=Meth_Cluster, color=Meth_Cluster)) + 
      scale_fill_manual(values = new_samp_info$colour_cat[[meth_var]]) + scale_color_manual(values = new_samp_info$colour_cat_border[[meth_var]]) + ylim(0, max(gene_set_rna_med$Gene_Set)+0.05*max(gene_set_rna_med$Gene_Set)) +
      labs(title=paste("Comparison of", gene_set, "Genes by Methylation Cluster"), y="Mean Log2(TMM+1)", x="") + 
      theme(panel.background=element_rect(fill='transparent'),
            axis.line = element_line(color="black"),
            axis.text.x.bottom = element_text(size=10), 
            axis.text = element_text(size=12),
            plot.title = element_text(size=14),
            legend.position = "none") +
      stat_compare_means(aes(group=Meth_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), comparisons=NULL)
  my.comparisons = list(c("Group1", "Group2"), c("Group1", "Group3"), c("Group1", "Group4"), c("Group2", "Group3"), c("Group2", "Group4"), c("Group3", "Group4"))
  p = p + stat_compare_means(method="wilcox.test", comparisons = my.comparisons, label.y = c(max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.08, 
                                                                                             max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.16, 
                                                                                             max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.24,
                                                                                             max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.32,
                                                                                             max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.40,
                                                                                             max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.48)) + 
    ylim(0, max(gene_set_rna_med$Gene_Set)+max(gene_set_rna_med$Gene_Set)*0.55)
  p = p + annotate("text", x=2, y=1, label=paste("Groups1&2 vs Group3 p = ", round(wilcox.test(gene_set_rna_med$Gene_Set[which(gene_set_rna_med$Meth_Cluster%in%c("Group1", "Group2"))], gene_set_rna_med$Gene_Set[which(gene_set_rna_med$Meth_Cluster%in%c("Group3"))])$p.value,4)))
  gene_set_comp_mean[[gene_set]] = ggplotGrob(p)   
}
gt = arrangeGrob(grobs=gene_set_comp_mean[c("NE", "non-NE")], ncol=2)
# *************************
# Fig 4 b & c
# *************************
grid.arrange(gt)

```


```{r look-at-meth-rna-relationships-for-genes, fig.height=5, fig.width=10 }
# meth_rna_genes = c(DLL3="treatmant", KDM1A="special", SLFN11="parpi", KDM1A="special", SEZ6="special", CD276="special", TACSTD2="special" )
meth_rna_genes = c(ASCL1="sclc", NEUROD1="sclc", POU2F3="sclc", YAP1="sclc", #KDM1A="special", EZH2="special", SLFN11="parpi",
                   MYC="oncogene", MYCN="oncogene", MYCL1="oncogene", REST="sclc", PLCG2="special" , HMGB3="extra", #DLL3="treatmant", SEZ6="special", CD276="special", TACSTD2="special", 
                   CD274="extra", CTLA4="extra" #, PDCD1="extra", NOTCH1="non-neuro", NOTCH2="non-neuro", NOTCH3="non-neuro"
                   )
# create CNV data for RNASeq samples and genes above
#
#  *************** copy_dir = <directory where copy number details are stored> *********************
CNV_data = NULL
for (donor in gsub("_T", "", colnames(RNASeq_TMM)[grep("_T", colnames(RNASeq_TMM))])){
  sample_study = dplyr::filter(study_data, donorLabel==donor, nchar(ascatAnalysis)>25)
  if (nrow(sample_study)>1 && length(unique(sample_study$testSequencingPlatform))>1){
    sample_study %<>% dplyr::filter(!testSequencingPlatform=="MGIseq2000")
  }
  sample.copy = paste0(copy_dir, donor, "/", gsub("analysis:", "", sample_study$ascatAnalysis), ".copynumber.ascat.tsv")
  if (file.exists(sample.copy)){
    message(paste("ascat copy number exists for", donor))
    sample.ascat = as.data.frame(fread(file=sample.copy))
    sample.gene = as.data.frame(fread(file=paste0(copy_dir, donor, "/", gsub("analysis:", "", sample_study$ascatAnalysis), ".genes.pc.best.cn.txt")))
    sample.gene %<>% dplyr::filter(gene_symbol%in%gsub("MYCL1", "MYCL", names(meth_rna_genes)))
    # message(paste("all genes are", paste(unique(sample.gene$SegmentDescription, collapse="or"))))
    sample.cn = data.frame(Sample_Name = rep(donor, length(names(meth_rna_genes))),
                           Genes = names(meth_rna_genes),
                           CN_values = sapply(gsub("MYCL1", "MYCL", names(meth_rna_genes)), function(x) ifelse(dplyr::filter(sample.gene, gene_symbol==x)$SegmentDescription=="span",
                                                                                                               (dplyr::filter(sample.ascat, chromosome==sample.gene$chromosome[which(sample.gene$gene_symbol==x)]) %>%
                                                                                                                      dplyr::filter(start<sample.gene$gene_start[which(sample.gene$gene_symbol==x)]&end>sample.gene$gene_end[which(sample.gene$gene_symbol==x)]))$median.rawCN,
                                                                                                               (dplyr::filter(sample.ascat, chromosome==sample.gene$chromosome[which(sample.gene$gene_symbol==x)]) %>%
                                                                                                                      dplyr::filter((start>sample.gene$gene_start[which(sample.gene$gene_symbol==x)]&start>sample.gene$gene_end[which(sample.gene$gene_symbol==x)])|
                                                                                                                                      (end>sample.gene$gene_start[which(sample.gene$gene_symbol==x)]&end<sample.gene$gene_end[which(sample.gene$gene_symbol==x)])|
                                                                                                                                      (start>sample.gene$gene_start[which(sample.gene$gene_symbol==x)]&end<sample.gene$gene_end[which(sample.gene$gene_symbol==x)])))$median.rawCN)))
    CNV_data = rbind(CNV_data, sample.cn)
  } else {
    message(paste("No ascat copy number analysis exists for", donor))
  }
}
meth_rna_gene_plots = list()
for (gene in names(meth_rna_genes)){
  meth_rna_gene_plots[[gene]] = draw.meth.rna.box.corr.plot(meth.data=meth_norm,
                                                            samp.info = new_samp_info$Sample_info,
                                                            RNA.info=RNASeq_TMM,
                                                            cnv.data = CNV_data,
                                                            cat.cols = new_samp_info$colour_cat,
                                                            cat.cols.bold = new_samp_info$colour_cat_border,
                                                            gene.info=gene_info,
                                                            genes=gene,
                                                            gene.type=meth_rna_genes[gene],
                                                            violin=FALSE,
                                                            plot.sd=0.1,
                                                            multi.test = "kruskal.test",
                                                            single.test = "wilcox.test",
                                                            all.comp = ifelse(gene%in%c("CD274", "CTLA4"), TRUE, FALSE))
}
meth_rna_gene_sets = c("apm", "cytolytic")
meth_rna_gene_set_plots = list()
for (gene_set in meth_rna_gene_sets){
  gene = genes_of_interest[[gene_set]]$Hugo_Symbol
  meth_rna_gene_set_plots[[gene_set]] = draw.meth.rna.box.corr.plot(meth.data=meth_norm,
                                                                    samp.info = new_samp_info$Sample_info,
                                                                    RNA.info=RNASeq_TMM,
                                                                    cat.cols = new_samp_info$colour_cat,
                                                                    cat.cols.bold = new_samp_info$colour_cat_border,
                                                                    gene.info=gene_info,
                                                                    genes=gene,
                                                                    gene.type=gene_set,
                                                                    violin=FALSE,
                                                                    plot.sd=0.1,
                                                                    multi.test = "kruskal.test",
                                                                    single.test = "wilcox.test",
                                                                    all.comp = TRUE)
}
```


```{r plot-meth-rna-corr-plots-for-paper-ascl1-neurod1-pou2f3-yap1, fig.height=20, fig.width=25}
# *************************
# Fig S3 a, b, c & d
# *************************
grid.arrange(meth_rna_gene_plots$ASCL1$meth$Meth_Promoter_Var, meth_rna_gene_plots$ASCL1$meth$Meth_Body_Var, meth_rna_gene_plots$ASCL1$rna, meth_rna_gene_plots$ASCL1$corr$Meth_Promoter_Var, meth_rna_gene_plots$ASCL1$corr$Meth_Body_Var,
             meth_rna_gene_plots$NEUROD1$meth$Meth_Promoter_Var, meth_rna_gene_plots$NEUROD1$meth$Meth_Body_Var, meth_rna_gene_plots$NEUROD1$rna, meth_rna_gene_plots$NEUROD1$corr$Meth_Promoter_Var, meth_rna_gene_plots$NEUROD1$corr$Meth_Body_Var,
             meth_rna_gene_plots$POU2F3$meth$Meth_Promoter_Var, meth_rna_gene_plots$POU2F3$meth$Meth_Body_Var, meth_rna_gene_plots$POU2F3$rna, meth_rna_gene_plots$POU2F3$corr$Meth_Promoter_Var, meth_rna_gene_plots$POU2F3$corr$Meth_Body_Var,
             meth_rna_gene_plots$YAP1$meth$Meth_Promoter_Var, meth_rna_gene_plots$YAP1$meth$Meth_Body_Var, meth_rna_gene_plots$YAP1$rna, meth_rna_gene_plots$YAP1$corr$Meth_Promoter_Var, meth_rna_gene_plots$YAP1$corr$Meth_Body_Var, nrow=4)

```



```{r plot-rna-plots-for-paper-amp-cytolytic-kdm1a-ezh2-pdl1-ctla4, fig.height=5, fig.width=25}
# *************************
# Fig 3 b, c, d, e & f
# *************************
grid.arrange(gp2_cells_p$`T-cells_CD8`, meth_rna_gene_set_plots$apm$rna, meth_rna_gene_set_plots$cytolytic$rna, meth_rna_gene_plots$CD274$rna, meth_rna_gene_plots$CTLA4$rna, nrow=1) 

```

```{r plot-rna-plots-for-paper-myc-mycn-mycl1-rest-plcg2-hmgb3, fig.height=5, fig.width=10}
# *************************
# Fig 4 d & e
# *************************
grid.arrange(meth_rna_gene_plots$PLCG2$rna, meth_rna_gene_plots$HMGB3$rna, nrow=1)

```


```{r plot-meth-rna-corr-plots-for-paper-myc-mycn-mycl1-rest-plcg2, fig.height=20, fig.width=30}
# grid and plots overview
gg_meth_rna_plots = list(meth_rna_gene_plots$MYC$rna, meth_rna_gene_plots$MYC$meth$Meth_Promoter, meth_rna_gene_plots$MYC$meth$Meth_Body_Var, meth_rna_gene_plots$MYC$corr$Meth_Promoter, meth_rna_gene_plots$MYC$corr$Meth_Body_Var, meth_rna_gene_plots$MYC$corr$CNV_value,
                         meth_rna_gene_plots$MYCN$rna, meth_rna_gene_plots$MYCN$meth$Meth_Promoter_Var, meth_rna_gene_plots$MYCN$meth$Meth_Body_Var, meth_rna_gene_plots$MYCN$corr$Meth_Promoter_Var, meth_rna_gene_plots$MYCN$corr$Meth_Body_Var, meth_rna_gene_plots$MYCN$corr$CNV_value,
                         meth_rna_gene_plots$MYCL1$rna, meth_rna_gene_plots$MYCL1$meth$Meth_Promoter, meth_rna_gene_plots$MYCL1$meth$Meth_Body_Var, meth_rna_gene_plots$MYCL1$corr$Meth_Promoter, meth_rna_gene_plots$MYCL1$corr$Meth_Body_Var, meth_rna_gene_plots$MYCL1$corr$CNV_value,
                         meth_rna_gene_plots$REST$rna, meth_rna_gene_plots$REST$meth$Meth_Promoter_Var, meth_rna_gene_plots$REST$meth$Meth_Body_Var, meth_rna_gene_plots$REST$corr$Meth_Promoter_Var, meth_rna_gene_plots$REST$corr$Meth_Body_Var, meth_rna_gene_plots$REST$corr$CNV_value)
                         # meth_rna_gene_plots$PLCG2$meth$Meth_Promoter_Var, meth_rna_gene_plots$PLCG2$meth$Meth_Body_Var, meth_rna_gene_plots$PLCG2$rna, meth_rna_gene_plots$PLCG2$corr$Meth_Promoter_Var, meth_rna_gene_plots$PLCG2$corr$Meth_Body_Var)
gt = arrangeGrob(grobs=gg_meth_rna_plots, layout_matrix=rbind(c(1, 2, 3, 4, 5, 6),
                                                              c(7, 8, 9, 10, 11, 12),
                                                              c(13, 14, 15, 16, 17, 18),
                                                              c(19, 20, 21, 22, 23, 24)))
                                                              # c(21, 22, 23, 24, 25)))
# *************************
# Fig S4 a, b, c & d
# *************************
grid.arrange(gt)

```


```{r compare-sclc-rna-with-other-subtypes, fig.width=12, fig.height=7  }
rna_genes = c(SEZ6="special", CD276="special", TACSTD2="special", DLL3="treatment", SLFN11="parpi")
rna_study_data %<>% dplyr::mutate(Subtype = sapply(Donor, function(x) ifelse(paste0(x, "_T")%in%colnames(RNASeq_TMM), 
                                                                             cluster_data$Meth_Cluster[which(cluster_data$Donor_Label==paste0(x, "_T"))],
                                                                             rna_study_data$`Tumour type`[which(rna_study_data$Donor==x)])))
gp_levels = c("Group1", "Group2", "Group3", "Group4", "Adeno", "NSCLC", "Squamous")
plot_rna_info = dplyr::filter(rna_study_data, Subtype%in%gp_levels) 
plot_rna_info$Subtype = factor(plot_rna_info$Subtype, levels=gp_levels)

violin=FALSE
multi.test = "kruskal.test"
single.test = "wilcox.test"
sub_col = c(new_samp_info$colour_cat_border$Meth_Cluster, sample_border_colours$Sample_Group[c("Adeno", "NSCLC", "Squamous")])
sub_fill = c(new_samp_info$colour_cat$Meth_Cluster, sample_colours$Sample_Group[c("Adeno", "NSCLC", "Squamous")])
exp.grob = list()
for (rna_gene in names(rna_genes)){
  plot_rna_info %<>% dplyr::mutate(RNA_Exp = sapply(Donor, function(x) log2(RNASeq_TMM_all[which(RNASeq_TMM_all$`Associated Gene Name`==rna_gene), paste0(x, "_T")]+1)))
  p.exp = ggplot(data=plot_rna_info, aes(x=Subtype, y=RNA_Exp))
    if (violin){
      p.exp = p.exp + geom_violin(width=0.9, aes(fill=Subtype, color=Subtype)) + 
        stat_summary(fun=mean, geom="point", size=2, color=sub_col[gp_levels])
    } else {
      p.exp = p.exp + geom_boxplot(aes(fill=Subtype, color=Subtype))
    }
    p.exp = p.exp + scale_fill_manual(values = sub_fill) + scale_color_manual(values = sub_col) + ylim(0, max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.55) + #ylim(0,max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.1) +
      labs(title=paste0(rna_gene, " Exp per Meth Cluster with other Lung Subtypes."), 
           y=paste(rna_gene, "Expresion\n", "Log2(exp+1)"), 
           x="") + 
      theme(panel.background=element_rect(fill='transparent'),
            axis.line = element_line(color="black"),
            axis.title = element_text(size=18),
            axis.text.x.bottom = element_text(size=14), # hjust=1, size=10, angle=90
            axis.text = element_text(size=12),
            plot.title = element_text(size=18),
            legend.position = "none") 
  my.comparisons = list(c("Group1", "Group2"), c("Group1", "Group3"), c("Group1", "Group4"), c("Group2", "Group3"), c("Group2", "Group4"), c("Group3", "Group4"))
  p.exp = p.exp + stat_compare_means(method="wilcox.test", comparisons = my.comparisons, label.y = c(max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.08, 
                                                                                             max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.16, 
                                                                                             max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.24,
                                                                                             max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.32,
                                                                                             max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.40,
                                                                                             max(plot_rna_info$RNA_Exp)+max(plot_rna_info$RNA_Exp)*0.48)) + 
    stat_compare_means(aes(group=Subtype), method=ifelse(length(gp_levels)>2, multi.test, single.test), label="p.format", label.x.npc = 0.7) + 
    stat_compare_means(aes(group=Subtype), method=ifelse(length(gp_levels)>2, multi.test, single.test), label="p.signif", label.x.npc = 0.85)
  if (rna_gene=="SLFN11"){
    p.exp = p.exp + annotate("text", x=2, y=1, label=paste("Groups1&2 vs Groups3&4   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1", "Group2"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group3", "Group4"))])$p.value,4)))
    p.exp = p.exp + annotate("text", x=2, y=0.5, label=paste("Groups1&3 vs Groups2&4   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1", "Group3"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group2", "Group4"))])$p.value,4)))
    # p.exp = p.exp + annotate("text", x=2, y=0, label=paste("Groups1 vs Groups2&4   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group2", "Group4"))])$p.value,4)))
  } else if (rna_gene=="DLL3"){
    p.exp = p.exp + annotate("text", x=2, y=1, label=paste("Groups1&2 vs Group3   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1", "Group2"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group3"))])$p.value,4)))
    p.exp = p.exp + annotate("text", x=2, y=0.5, label=paste("Groups1&2 vs Group4   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1", "Group2"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group4"))])$p.value,4)))
     p.exp = p.exp + annotate("text", x=2, y=0, label=paste("Groups1&2 vs Groups3&4   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1", "Group2"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group3", "Group4"))])$p.value,4)))
  } else if (rna_gene=="TACSTD2"){
    p.exp = p.exp + annotate("text", x=2, y=0, label=paste("Groups1&2&3 vs Group4   p = ", round(wilcox.test(plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group1", "Group2", "Group3"))], plot_rna_info$RNA_Exp[which(plot_rna_info$Subtype%in%c("Group4"))])$p.value,4)))
  }
  exp.grob[[rna_gene]] = ggplotGrob(p.exp)
}


```

```{r plot-rnaseq_comparison-plots-for-paper-sez6-tacstd2-dll3-cd276-slfn11-ezh2-kdm1a, fig.height=20, fig.width=18}
gg_rna_lung = list(exp.grob$SLFN11, exp.grob$CD276, exp.grob$SEZ6, exp.grob$DLL3, exp.grob$TACSTD2) # , exp.grob$EZH2, exp.grob$KDM1A
gt = arrangeGrob(grobs=gg_rna_lung, layout_matrix=rbind(c(1, 1, 1, 1, 1, 1, NA, NA, NA, 2, 2, 2, 2, 2, 2),
                                                        c(3, 3, 3, 3, 3, 3, NA, NA, NA, 4, 4, 4, 4, 4, 4),
                                                        c(NA, NA, NA, NA, 5, 5, 5, 5, 5, 5, NA, NA, NA, NA, NA)))
# *************************
# Fig 5 a, b, c, d, e
# *************************
grid.arrange(gt)

```

