---
title: |
  | \vspace*{3cm}  Fielding Lung Illumina Sequencing 
  | Summary for WGS - SCLC 
  | Variant Landscape for Driver Genes \vspace*{3cm}
author: |
  | Vanessa Lakis \vspace{3cm}  
date: |
  | `r format(Sys.time(), "%d %B, %Y")` \newpage
output: 
  pdf_document:
    latex_engine: xelatex
    number_sections: yes
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_tex: yes
  html_document:
    toc: true
highlight: zenburn
  
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{xcolor, hyperref}
  - \usepackage{lmodern}
  - \usepackage{courier}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{pdflscape}  
  - \newcommand{\blandscape}{\begin{landscape}} 
  - \newcommand{\elandscape}{\end{landscape}}
  - \setlength{\headheight}{36.5pt}
  - \setlength{\footskip}{28pt}
  - \pagestyle{fancy}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \cfoot{\thepage}
  - \rhead{\scriptsize The University of Queensland \\ 280-284 Sir Fred Schonell Dr, St Lucia, QLD 4067 Australia\\T +61 (7) 3365 1111 | W \url{www.uq.edu.au}}
  - \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \usepackage{float}
---

```{r, global-options, include=FALSE}
# ***************
# load packages required
# ***************
library(knitr)
knitr::opts_chunk$set(dpi = 100, echo=FALSE, warning=FALSE, message=FALSE, 
                      fig.show='asis', fig.keep = 'all', fig.pos="H", fig.align='center', fig.width=6, fig.height=6) 
# to remove message from xtable
library(xtable)
options(xtable.comment = FALSE)

# acquire library for the analysis
packages = c("DT", "data.table", "dplyr", "tidyr", "magrittr", "readr",
             "ggalt", "ggplot2", "gtable", "scales", "gridExtra", "ggpubr", "ggfortify", 
             "RColorBrewer", "ComplexHeatmap", "circlize", "stringi", "reshape2", "maftools", "cowplot", "copynumber")
loadpak = function(pkg){
  sapply(pkg, require, character.only=TRUE)
}
loadpak(packages)

```


```{r, import-declarations-and-functions, include=FALSE}
# **********
# import functions for script
# *********
#
#   ********************** nominate study directory ***************************
#
study_dir = ""
source(paste0(study_dir, "GeneLandscape_Summary/WGS_SCLC_Analysis/GeneLandscape_Summary_Fielding_-_Lung_WGS_SCLC_Functions_Upload.R"))
# ********** 
# import declarations for script
# *********
source(paste0(study_dir, "GeneLandscape_Summary/WGS_SCLC_Analysis/GeneLandscape_Summary_Fielding_-_Lung_Declarations_Upload.R"))


```

\newpage  

```{r, read-in-dataframes}
if (full_study){
  all_samples=c("WGS", "WES")
} else if (WGS_only){
  all_samples = "WGS"
} else if (WES_only){
  all_samples = "WES"
} else {
  all_samples = NULL
}
SNV = list(WGS=NULL,
           WES=NULL)
CNV = list(WGS=NULL,
           WES=NULL)
SV = list(WGS=NULL,
          WES=NULL)
for(samp_type in all_samples){
  RData_file = paste(working_dir, "WGS_SCLC_Analysis/RData/", study_prefix, 
                     ifelse(VEP_anno, "_VEP", "_MAF"), 
                     ifelse(samp_type%in%c("WGS", "WES"), 
                            paste("_", nrow(study_data[grep(samp_type, study_data$UniqueLabel),]), "_", samp_type, sep=""), 
                            ""), "_filtered", ".RData", sep="")
  if (samp_type%in%c("WGS", "WES")){
    if (!file.exists(RData_file)){
      SNV_data = populate.snv(study.data=study_data[grep(samp_type, study_data$UniqueLabel),])
      SV_data = populate.sv(study.data=study_data[grep(samp_type, study_data$UniqueLabel),])
      CNV_data = populate.cnv(study.data=study_data[grep(samp_type, study_data$UniqueLabel),], samp.type = samp_type)
      if (samp_type=="WGS"){
        wgs_study_data = study_data[grep(samp_type, study_data$UniqueLabel),]
        save(wgs_study_data, SNV_data, SV_data, CNV_data, file=RData_file)
      } else if (samp_type=="WES"){
        wes_study_data = study_data[grep(samp_type, study_data$UniqueLabel),]
        save(wes_study_data, SNV_data, SV_data, CNV_data, file=RData_file)
      }
    } else {
      load(RData_file)
    }
  } 
  SNV[[samp_type]] = SNV_data
  CNV[[samp_type]] = CNV_data
  SV[[samp_type]] = SV_data
}

```


```{r, create-data-for-analysis}
# create the data and sample data for the analysis
SNV_data = NULL
CNV_data = NULL
SV_data = NULL
snv_columns = colnames(SNV[[samp_type]]$data)

for (samp_type in all_samples){
  if(length(all_samples) == 1){
    SNV_data = SNV[[samp_type]]$data
    if (!is.null(CNV[[samp_type]])){
      CNV_data = CNV[[samp_type]]
    }
    if (!is.null(SV[[samp_type]]$data)){
      SV_data = SV[[samp_type]]$data
    }
  } else {
    SNV_data = rbind(SNV_data, dplyr::select(SNV[[samp_type]]$data, all_of(snv_columns)))
    if (samp_type %in% c("WGS", "WES") && !is.null(CNV[[samp_type]])){
      CNV_data = rbind(CNV_data,
                       data.frame(DonorLabel = CNV[[samp_type]]$DonorLabel,
                                  UniqSampleLabel = CNV[[samp_type]]$UniqSampleLabel,
                                  Hugo_Symbol = CNV[[samp_type]]$Hugo_Symbol,
                                  Copy_Number_Variant = CNV[[samp_type]]$Copy_Number_Variant,
                                  Copynumber = CNV[[samp_type]]$Copynumber))
      }
    if (!is.null(SV[[samp_type]]$data)){
      SV_data = SV[[samp_type]]$data
    }
  }
}

```

\newpage

# `r paste("Summarise Mutational Landscape for", length(Gene_list$Hugo_Symbol), "Supplied Genes", sep=" ") `.

This section will summarise the mutational landscape of a set of genes.  Where a gene has more than one mutation called the mutation calls are concatenated or represented as a multi-hit.  
Two plots are presented:  

1. **Single Nucleotide Variant Landscape** - This plot is designed to report either the maf annotations of nonsense, frameshift (insertions and deletions), missense mutations, in-frame (insertions and deletions), splice site and translation start site mutations or if flagged the vep annotations of stop gained, frameshift (insertions and deletions), missense mutations, in-frame (insertions and deletions), splice site mutations (acceptor, donor of region), start lost and stop lost.  
2. **Full Variant Landscape** - This plot combines the results of the first plot with any copy number alteration or structural variant loss of function.  If the flags are set as only loss or only gain for Copy number changes then the report will represent cnLOH as a loss and only high gains will be reported.  If a gene has more than one mutation then this plot will only report the mutation of highest consequence.  

\newpage

```{r, prepare-data-for-summary-mapping-data-frames, results='asis'}
if (!first_pass){
  # prepare the data
  if (!is.null(SNV_data)){
    SNV_data = prepare.data(df=as.data.frame(SNV_data),
                            type="SNV",
                            mut.types=SNV_mutations,
                            variant.class="Plot_Class")
  }
  if (!is.null(SV_data)){
    SV_data = prepare.data(df=as.data.frame(SV_data),
                           type="SV",
                           mut.types=SV_mutations,
                           variant.class="likely_consequence")
    if (!is.null(SV_data)){
      SV_data %<>% mutate(Reported_consequence= ifelse(grepl("Loss_of_Function", SV_data$likely_consequence),
                                                       "Loss_of_Function",
                                                       ifelse(grepl("Intragene_Fusion", SV_data$likely_consequence),
                                                              "Intragene_Fusion",
                                                              ifelse(grepl("Intra_intron", SV_data$likely_consequence),
                                                                     "Intra_intron",
                                                                     ifelse(grepl("Intergene_Fusion", SV_data$likely_consequence),
                                                                            "Intergene_Fusion", "BadLuck")))))
    }
  }
  if (!is.null(CNV_data)){
    CNV_data = prepare.data(df=as.data.frame(CNV_data),
                            type="CNV",
                            mut.types=CNV_mutations,
                            variant.class="Copy_Number_Variant")
  }
}

```


```{r, create-data-for-summary-mapping-data-frames, results='asis'}
plot_types = c("full") # all sample subtypes
if (length(all_samples)>=1){
  att = unique(c("donorLabel", "UniqueLabel", "Donor_class", colnames(study_data)[which(colnames(study_data)%in% top_annots)]))
  full_study_data = select(study_data, all_of(att))
  if (sample_order){
    sample_levels %<>% dplyr::filter(donorLabel %in% study_data$donorLabel) 
    if (!samp_id %in% colnames(sample_levels)){
      sample_levels %<>% dplyr::mutate(sampID = sapply(donorLabel, function(x) study_data[which(study_data$donorLabel==x), samp_id]))
      colnames(sample_levels)[which(colnames(sample_levels)=="sampID")] = samp_id
    }
    full_study_data[,samp_id]= factor(full_study_data[,samp_id], levels=c(sample_levels[,samp_id]))
    full_study_data = full_study_data[order(full_study_data[,samp_id]),]
  } else {
    full_study_data %<>% arrange(Donor_class, desc(Cellularity), desc(UniqueLabel))
    # create an index to arrange samples by cellularity high to low within status by number of donors, 
    full_study_data$index = vector(mode="integer", length=nrow(full_study_data))
    donor_list = unique(study_data$donorLabel)
    for (donor in donor_list){
      full_study_data$index[c(grep(donor, full_study_data$UniqueLabel))] = length(grep(donor, full_study_data$UniqueLabel))
    }
    full_study_data %<>% arrange(Donor_class, desc(index))
    full_study_data$donorLabel = factor(full_study_data$donorLabel, levels=unique(full_study_data$donorLabel))
  }
  donor_list = list()
  plot_study_data = list()
  for (plot in plot_types){
    if (plot=="full"){
      donor_list[[plot]] = full_study_data$UniqueLabel
      plot_study_data[[plot]] = full_study_data
    } 
  }
}


```


```{r, create-summary-mapping-data-frames, results='asis'}
if (!first_pass){
  # create mutational landscape data
  if (!is.null(SNV_data)){
    SNV_landscape=list()
    SNV_Effects=list()
    for (plot in plot_types){
      SNV_mut_data = create.mut.landscape(type="SNV",
                                          df.list=list(SNV = SNV_data),
                                          donor.list=donor_list[[plot]],
                                          study.data=plot_study_data[[plot]],
                                          top.annos = top_annots)
      SNV_landscape[[plot]] = SNV_mut_data$landscape
      SNV_Effects[[plot]] = unlist(unique(SNV_mut_data$effects))
    }
  } else {
    SNV_landscape = NULL
  }
  if (!is.null(CNV_data) || !is.null(SV_data)){
    mut_landscape = list()
    Variant_Effects = list()
    for (plot in plot_types){
      mut_data = create.mut.landscape(type="all",
                                      df.list=list(SNV = SNV_data, SV = SV_data[grep("Loss_of_Function", SV_data$Reported_consequence),], CNV = CNV_data), 
                                      donor.list=donor_list[[plot]],
                                      study.data=plot_study_data[[plot]],
                                      top.annos = top_annots)
      mut_landscape[[plot]] = mut_data$landscape
      Variant_Effects[[plot]] = unlist(unique(mut_data$effects))
    }
  } else {
    mut_landscape = NULL
  }
  if (!is.null(top_genes)){
    if (!is.null(SNV_data) || !is.null(CNV_data) || !is.null(SV_data)){
      top_mut_data = list()
      top_effects = list()
      if (add_ctdna){
        top_study_data = plot_study_data[[plot]]
        top_cf_annots = top_annots
      }
      for (plot in plot_types){
        top_data = create.mut.landscape(type="TOP",
                                        df.list=list(SNV = SNV_data, SV = SV_data[grep("Loss_of_Function", SV_data$Reported_consequence),], CNV = CNV_data), 
                                        donor.list=donor_list[[plot]],
                                        study.data=top_study_data,
                                        top.annos = top_cf_annots,
                                        top.genes = top_genes,
                                        add.ctdna = add_ctdna)
        top_mut_data[[plot]] = top_data$landscape
        top_effects[[plot]] = unlist(unique(top_data$effects))
      }
    } else {
      top_genes = NULL
    }
  }
}

```

```{r, create-summary-data-matrix-for-plotting}
# ***************
# create the data and parameters for plotting matrix
# ***************
figure_order = arrange(Genes_of_interest, desc(SNV)) 
if (!first_pass){
  # populate data for SNV plots
  if (!is.null(SNV_landscape)){
    if (specific_gene_order){
      gene_order_file = SNV_gene_order_file
      gene_order = as.data.frame(fread(file=gene_order_file))
      gene_order = gene_order$GeneID
    } else {
      gene_order = figure_order$Hugo_Symbol # NULL
    }
    SNV_plots = list()
    for (plot in plot_types){
      SNV_plots[[plot]] = get.plot.data(type="SNV",
                                        v.effects=SNV_Effects[[plot]],
                                        v.df=as.data.frame(SNV_landscape[[plot]]),
                                        v.filename="SNV_matrix",
                                        data=SNV_data,
                                        donor.list=donor_list[[plot]],
                                        study.data = plot_study_data[[plot]],
                                        plot.type = plot,
                                        gene.order=gene_order)
    }
  } else {
    SNV_plots = NULL
  }
  # populate data for summary plots i.e. a combination of SNV, CNV and SV data
  if (!is.null(mut_landscape)){
    data_summ = SNV_data
    if (specific_gene_order){
      gene_order_file = full_gene_order_file
      gene_order = as.data.frame(fread(file=gene_order_file))
      gene_order = gene_order$GeneID
    } else {
      gene_order = figure_order$Hugo_Symbol # NULL
    }
    summary_plots = list()
    for (plot in plot_types){
      summary_plots[[plot]] = get.plot.data(type="summary",
                                            v.effects=Variant_Effects[[plot]],
                                            v.df=as.data.frame(mut_landscape[[plot]]),
                                            v.filename="Summary_matrix",
                                            data=data_summ,
                                            donor.list=donor_list[[plot]],
                                            study.data = plot_study_data[[plot]],
                                            plot.type = plot,
                                            gene.order=gene_order)
    }
  } else {
    summary_plots = NULL
  }
  # populate data for top genes plots
  if (!is.null(top_genes)){
    gene_order = NULL
    if (add_ctdna) {
      top_gene_list = data.frame(Hugo_Symbol = c(top_genes, paste0(top_genes, "_ct")),
                                 SNV = sapply(rep(top_genes, 2), function(x) Genes_of_interest$SNV[which(Genes_of_interest$Hugo_Symbol==x)]),
                                 CNV = sapply(rep(top_genes, 2), function(x) Genes_of_interest$CNV[which(Genes_of_interest$Hugo_Symbol==x)]),
                                 SV = sapply(rep(top_genes, 2), function(x) Genes_of_interest$SV[which(Genes_of_interest$Hugo_Symbol==x)]),
                                 Group = rep(top_genes, 2))
      top_gene_list$Group = factor(top_gene_list$Group, levels=top_genes)
      donor_list[[plot]] = donor_list[[plot]][which(donor_list[[plot]]%in%colnames(top_mut_data[[plot]]))]
    } else {
      top_gene_list = rbind(dplyr::select(Genes_of_interest, Hugo_Symbol, SNV, CNV, SV), 
                            data.frame(Hugo_Symbol=c("TP53_LOH", "RB1_LOH"),
                                       SNV = c(Genes_of_interest$SNV[which(Genes_of_interest$Hugo_Symbol=="TP53")], Genes_of_interest$SNV[which(Genes_of_interest$Hugo_Symbol=="RB1")]),
                                       CNV = c(Genes_of_interest$CNV[which(Genes_of_interest$Hugo_Symbol=="TP53")], Genes_of_interest$CNV[which(Genes_of_interest$Hugo_Symbol=="RB1")]),
                                       SV = c(Genes_of_interest$SV[which(Genes_of_interest$Hugo_Symbol=="TP53")], Genes_of_interest$SV[which(Genes_of_interest$Hugo_Symbol=="RB1")]))
                            )
    }
    top_gene_list %<>% dplyr::arrange(desc(SNV))
    gene_order = top_gene_list$Hugo_Symbol
    TOP_plots = list()
    for (plot in plot_types){
      TOP_plots[[plot]] = get.plot.data(type="TOP",
                                        v.effects=top_effects[[plot]],
                                        v.df=as.data.frame(top_mut_data[[plot]]),
                                        v.filename="TOP_matrix",
                                        data=data_summ,
                                        donor.list=donor_list[[plot]],
                                        gene.list=top_gene_list,
                                        study.data = top_study_data,
                                        plot.type = plot,
                                        gene.order=gene_order)
    }
  } else {
    TOP_plots = NULL
  }
}
```


```{r, organise-colours-for-plots}
# ***************
# set up colours to be used in the plots and to populate values
# ***************
if (!first_pass){
  if (!is.null(summary_plots)){
    summary_all_colours = list()
    summary_colours = list()
    for (plot in plot_types){
      summary_all_colours[[plot]] = get.plot.colours(plot.data = summary_plots[[plot]],
                                                     gene.list = figure_order, 
                                                     type="summary_all")
    }
  } else {
    summary_colours = NULL
    summary_all_colours = NULL
  }
  if (!is.null(SNV_plots)){
    SNV_colours = list()
    for (plot in plot_types){
      SNV_colours[[plot]] = get.plot.colours(plot.data=SNV_plots[[plot]], 
                                             gene.list = figure_order, 
                                             type="SNV")
    }
  } else {
    SNV_colours = NULL
  }
  if (!is.null(TOP_plots)){
    TOP_colours = list()
    for (plot in plot_types){
      TOP_colours[[plot]] = get.plot.colours(plot.data=TOP_plots[[plot]], 
                                             type="TOP")
    }
  } else {
    TOP_colours = NULL
  }
}

```


```{r, functions-for-plot-matrices-and-legends}
if (!first_pass){
  # ***************
  # set up the parameters for the plots
  # ***************
  plot_type = c("SNV", "summary_all", "TOP")
  plot_data = list()
  plot_mutations = list()
  plot_titles = list()
  plot_colours = list()
  var_effects = list()
  for (plot in plot_types){
    plot_data[[plot]] = list(SNV_plots[[plot]]$Data, summary_plots[[plot]]$Data, TOP_plots[[plot]]$Data)
    names(plot_data[[plot]]) = plot_type
    plot_mutations[[plot]] = list(SNV_plots[[plot]]$NumMutations, summary_plots[[plot]]$NumMutations, TOP_plots[[plot]]$NumMutations)
    names(plot_mutations[[plot]]) = plot_type
    plot_titles[[plot]] = c("Single Nucleotide", "Full", "Top Genes")
    names(plot_titles[[plot]]) = plot_type
    plot_colours[[plot]] = list(SNV_colours[[plot]], summary_all_colours[[plot]], TOP_colours[[plot]])
    names(plot_colours[[plot]]) = plot_type
    var_effects[[plot]] = list(SNV_plots[[plot]]$Effects, summary_plots[[plot]]$Effects, TOP_plots[[plot]]$Effects)
    names(var_effects[[plot]]) = plot_type
  }
}

```

\blandscape  

## Single Nucleotide Variant Landscape for `r study_name `.  

```{r, plot-SNV-matrix-1, fig.width=16, fig.height=16, fig.align='center', results='asis'}
# single nucleotide variant plots
 if (!first_pass & is.null(top_genes)){
   for (plot_type in plot_types){
    if (!is.null(var_effects[[plot_type]][["SNV"]])){
      SNV_plot = create.matrix.plot(type="SNV",
                                    plot.data=plot_data[[plot_type]],
                                    plot.mut = plot_mutations[[plot_type]],
                                    plot.colours=plot_colours[[plot_type]],
                                    plot.titles=plot_titles[[plot_type]],
                                    study.data = plot_study_data[[plot_type]],
                                    gene.list = figure_order)
      if (!is.null(SNV_plot$gene.plot) && !is.null(SNV_plot$anno.plot) && !is.null(SNV_plot$tmb.plot)){
        grid.arrange(plot_grid(SNV_plot$tmb.plot, NULL, SNV_plot$anno.plot, NULL, SNV_plot$gene.plot, ncol = 1, align="v", rel_heights = SNV_plot$plot.heights))
      } else if (!is.null(SNV_plot$gene.plot) && !is.null(SNV_plot$anno.plot)){
        grid.arrange(plot_grid(SNV_plot$anno.plot, NULL, SNV_plot$gene.plot, ncol = 1, align="v", rel_heights = SNV_plot$plot.heights))
      }
      cat("\\pagebreak  ")
    } else {
      cat(paste("No Single Nucleotide Variants have been called", plot_types[num], "samples for the", study_name, "study.\n", sep=" ") 
          )
      cat("\\pagebreak  ")
    }
  }
}

```

\elandscape  


```{r, plot-SNV-legend-1, fig.width=9, fig.height=6, fig.keep='last', fig.align='center', results='asis'}
# single nucleotide variant legends
if (!first_pass & is.null(top_genes)){
  for (plot_type in plot_types){
    if (!is.null(var_effects[[plot_type]][["SNV"]])){
      create.legend(type="SNV",
                    plot.colours=plot_colours[[plot_type]],
                    var.effects=var_effects[[plot_type]])
      cat("\\pagebreak  ")
    } 
  }
}

```


\blandscape  

## Full Variant Landscape for `r study_name `.  

```{r, plot-summary_all-matrix-1, fig.width=12, fig.height=12, fig.align='center', results='asis'}
# summary variant plots including SNV, CNV and SV
if (!first_pass & is.null(top_genes)){
  for (plot_type in plot_types){
    if (!is.null(var_effects[[plot_type]][["summary_all"]])){
      summary_all_plot = create.matrix.plot(type="summary_all",
                                            plot.data=plot_data[[plot_type]],
                                            plot.mut=plot_mutations[[plot_type]],
                                            plot.colours=plot_colours[[plot_type]],
                                            plot.titles=plot_titles[[plot_type]],
                                            study.data = plot_study_data[[plot_type]],
                                            gene.list=Genes_of_interest)
      if (!is.null(summary_all_plot$gene.plot) && !is.null(summary_all_plot$anno.plot) && !is.null(summary_all_plot$tmb.plot)){
        grid.arrange(plot_grid(summary_all_plot$tmb.plot, NULL, summary_all_plot$anno, NULL, summary_all_plot$gene.plot, ncol = 1, align="v", rel_heights = summary_all_plot$plot.heights))
      } else if (!is.null(summary_all_plot$gene.plot) && !is.null(summary_all_plot$anno.plot)){
        if (class(summary_all_plot$gene.plot)=="list") {
          # *************************
          # Fig S5 a (with no top_genes and add_ctDNA=FALSE)
          # *************************
          grid.arrange(plot_grid(summary_all_plot$anno.plot, NULL,
                                 summary_all_plot$gene.plot$Cell_Cycle_Regulation, NULL,
                                 summary_all_plot$gene.plot$Transcriptional_Regulation, NULL,
                                 summary_all_plot$gene.plot$Receptor_Kinase_P13K_signalling, NULL,
                                 summary_all_plot$gene.plot$Notch_signalling_NE_differentiation, NULL,
                                 summary_all_plot$gene.plot$Other, NULL, ncol = 1, align="v", rel_heights = summary_all_plot$plot.heights),
                       plot_grid(NULL, NULL,
                                 summary_all_plot$gene.hist$Cell_Cycle_Regulation, NULL,
                                 summary_all_plot$gene.hist$Transcriptional_Regulation, NULL,
                                 summary_all_plot$gene.hist$Receptor_Kinase_P13K_signalling, NULL,
                                 summary_all_plot$gene.hist$Notch_signalling_NE_differentiation, NULL,
                                 summary_all_plot$gene.hist$Other, NULL, ncol = 1, align="v", rel_heights = c(summary_all_plot$plot.heights[1:10], summary_all_plot$plot.heights[11]-4.8, summary_all_plot$plot.heights[12]+4.8)), layout_matrix=rbind(c(1,1,1,1,1,1,1,1,1,2)))
        } else {
          grid.arrange(plot_grid(summary_all_plot$anno.plot, NULL, summary_all_plot$gene.plot, NULL, ncol = 1, align="v", rel_heights = summary_all_plot$plot.heights))
        }
      }
      cat("\\pagebreak  ")
    } else {
      cat(paste("No Summary available for", plot_types[num], "samples for the", study_name, "study.\n", sep=" ") 
          )
      cat("\\pagebreak")
    }
  }
}

```

\elandscape  

```{r, plot-summary_all-legend-1, fig.width=9, fig.height=6, fig.keep='last', fig.align='center', results='asis'}
# summary variant legend for combined SNV, CNV and SV
 if (!first_pass & is.null(top_genes)){
   for (plot_type in plot_types) {
    if (!is.null(var_effects[[plot_type]][["summary_all"]])){
      # *************************
      # Legend S5 a
      # *************************
      create.legend(type="summary_all",
                    plot.colours=plot_colours[[plot_type]],
                    var.effects=var_effects[[plot_type]])
      cat("\\pagebreak  ")
    } 
  }
}

```

\blandscape  


```{r, plot-summary_top-matrix-1, fig.width=12, fig.height=7, fig.align='center', results='asis'}
# top genes variant plots
if (!first_pass & !is.null(top_genes)){
  for(plot_type in plot_types){
    if (!is.null(var_effects[[plot_type]][["TOP"]])){
      top_plot = create.matrix.plot(type="TOP",
                                    plot.data=plot_data[[plot_type]],
                                    plot.colours=plot_colours[[plot_type]],
                                    plot.titles=plot_titles[[plot_type]],
                                    study.data = top_study_data[which(top_study_data$UniqueLabel%in%unique(plot_data[[plot_type]]$TOP$DonorID)),],
                                    top.annots = top_cf_annots,
                                    gene.list = top_gene_list)
      if (!is.null(top_plot$gene.plot) && !is.null(top_plot$anno.plot) && !is.null(top_plot$tmb.plot)){
        grid.arrange(plot_grid(top_plot$tmb.plot, NULL, top_plot$anno, NULL, top_plot$gene.plot, ncol = 1, align="v", rel_heights = top_plot$plot.heights))
      } else if (!is.null(top_plot$gene.plot) && !is.null(top_plot$anno.plot)){
        if (class(top_plot$gene.plot)[1]=="list") {
          # *************************
          # Fig S8 (with top_genes and add_ctDNA=FALSE)
          # Fig S9 (with top_genes and add_ctDNA=TRUE)
          # *************************
          grid.arrange(plot_grid(top_plot$anno.plot, NULL,
                                 top_plot$gene.plot$TP53, NULL,
                                 top_plot$gene.plot$RB1, NULL, ncol = 1, align="v", rel_heights = top_plot$plot.heights), ncol=1)
        } else {
          grid.arrange(plot_grid(top_plot$anno.plot, NULL, top_plot$gene.plot, ncol = 1, align="v", rel_heights = top_plot$plot.heights))
          # ggsave2(filename = paste0(study_prefix, "_Summary_Top_Mutations_w_", ifelse(add_ctdna, "ctDNA_67Samples", "CN_LOH"), "_FINAL.pdf"),
          #     plot = grid.arrange(plot_grid(top_plot$anno.plot, NULL, top_plot$gene.plot, ncol = 1, align="v", rel_heights = top_plot$plot.heights)),
          #     device="pdf",
          #     path=paste0(results_dir, "Plots/Paper_Figures/"),
          #     width=12, height=7, units="in")
        }
      }
      cat("\\pagebreak  ")
      
    } else {
      cat(paste("No Summary available for", plot_types[num], "samples for the", study_name, "study.\n", sep=" ") 
          )
      cat("\\pagebreak")
    }
  }
}


```

\elandscape  

```{r, plot-top-legend-1, fig.width=9, fig.height=6, fig.keep='last', fig.align='center', results='asis'}
# top genes variant legends
if (!first_pass & !is.null(top_genes)){
  for (plot_type in plot_types){
    if (!is.null(var_effects[[plot_type]][["TOP"]])){
      # *************************
      # Legends S8 & S9
      # *************************
      create.legend(type="TOP",
                    plot.colours=plot_colours[[plot_type]],
                    var.effects=var_effects[[plot_type]])
      cat("\\pagebreak  ")
    }
  } 
}

```


```{r plot-genomic-information-by-cluster, fig.show='asis', fig.keep = 'all', fig.width=15, fig.height=10}
telomere_info = as.data.frame(fread(file=paste0(working_dir, "Spreadsheets/Fielding_-_Lung_Small_cell_Lung_cancer_qmotif.tsv")))
genomic_info = data.frame(DonorLabel = study_data$donorLabel,
                          Meth_Cluster = gsub("T", "Group", study_data$Meth_Cluster),
                          Ploidy = study_data$ascatPloidyScore,
                          TMB = study_data$TMB_nonSyn,
                          Log2_Telomere_Ratio = sapply(study_data$donorLabel, function(x) telomere_info$log2.telomere.ratio[which(telomere_info$donorLabel==x)]),
                          Total_Num_SV = sapply(study_data$donorLabel, function(x) SV$WGS$stats$SV_num[grep(x, SV$WGS$stats$UniqueLabel)])
                          ) 
genomic_info$Meth_Cluster = factor(genomic_info$Meth_Cluster, levels=c("Group1", "Group2", "Group3", "Group4"))
groups = c("Methylation Sub-Groups")
genomic_tests = c("TMB", "Total_Num_SV", "Ploidy", "Log2_Telomere_Ratio")
gene_clust_col = c(`Group1`="#49b05c", `Group2`="#5108b4", `Group3`="#f58919", `Group4`="#d9c80d")
gene_clust_fill = c(`Group1`="#b9e4c1", `Group2`="#d3c3ef", `Group3`="#f8d9b0", `Group4`="#edeac4")
sep_plots = list()
for (gt in genomic_tests){
  gp_levels = unique(study_data$Meth_Cluster)[order(unique(study_data$Meth_Cluster))]
  genomic_info %<>% dplyr::mutate(Y_var = genomic_info[,gt])
  p = ggplot(data=genomic_info, aes(x=Meth_Cluster, y=Y_var)) + geom_boxplot(aes(fill=Meth_Cluster, color=Meth_Cluster))  +  
    # geom_violin(width=0.9, aes(fill=Meth_Cluster, color=Meth_Cluster)) + stat_summary(fun=mean, geom="point", size=2, color=gene_clust_col[2:5]) + # mult=0.5, 
    scale_fill_manual(values = gene_clust_fill) + scale_color_manual(values = gene_clust_col) + ylim(min(genomic_info$Y_var), max(genomic_info$Y_var)+0.1*max(genomic_info$Y_var)) + 
      labs(title=paste(gsub("_", " ", gsub("Num", "#", gt)), "by Sub-Groups"), y=paste(gsub("_", " ", gsub("Num", "#", gt))), x=gsub("_", " ", groups)) + 
      theme(panel.background=element_rect(fill='transparent'),
            axis.line = element_line(color="black"),
            panel.grid.major.y = element_line(color="#dddddd", linewidth=0.25),
            # axis.text.x.bottom = element_text(hjust=1, size=12), #, angle=90
            axis.text = element_text(size=12),
            axis.title = element_text(size=18),
            plot.title = element_text(size=18),
            legend.position = "none") +
      stat_compare_means(aes(group=Meth_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), label="p.format", label.x.npc = "center") +  # other methods = t.test & anova for parametric test
      stat_compare_means(aes(group=Meth_Cluster), method=ifelse(length(gp_levels)>2, "kruskal.test", "wilcox.test"), label="p.signif", label.x.npc = 0.9)
  sep_plots[[gt]] = ggplotGrob(p)   
}
gg_gt_plots = list(sep_plots$Total_Num_SV, sep_plots$Ploidy, sep_plots$Log2_Telomere_Ratio, sep_plots$TMB)
gt_grob = arrangeGrob(grobs=gg_gt_plots, layout_matrix=rbind(c(1, 2, 3),
                                                              c(4, NA, NA)))


# *************************
# Fig S5 b, c, d & g
# *************************
grid.arrange(gt_grob)

```


